An integrated human forebrain organoid reveals microglia-mediated CD8+ T cell recruitment and neuroimmune dysfunction in Alzheimer’s disease pathology

Version 1

Abstract

Genetic evidence implicates immune dysfunction in Alzheimer’s disease (AD), yet human-specific neuroimmune mechanisms remain poorly defined. Here, we establish a modular human forebrain organoid platform that systematically integrates iPSC-derived microglia and CD8+ T cells to reconstitute multicellular Alzheimer’s disease pathology. This system enables functional interrogation of both innate and adaptive immune components in a human-relevant context. Using this platform, we demonstrate that microglia mediate amyloid-β clearance and neuronal maturation but also drive inflammatory activation and recruit CD8+ T cells through CCL4/5-CXCL10 signaling via CCR1/5 and CXCR3, establishing a neuroinflammatory feedback loop. Pharmacological targeting of CCR5 or CXCR3 blocks T cell recruitment and modulates autophagy in a microglia-dependent manner. This modular organoid platform provides a versatile tool for dissecting neuron–immune interactions and enables cell-type-specific therapeutic screening in human neuroinflammatory disease models.

Introduction

Alzheimer’s disease (AD) is featured by amyloid-beta (Aβ) accumulation, which triggers a complex cascade of glial activation, neuroinflammation, and neuronal damage (1,2). While resident microglia and astrocytes drive much of this local neuroinflammatory landscape (4,5,8,9,10), emerging evidence highlights the active infiltration of adaptive immune cells - particularly CD8+ T cells - into the AD brain parenchyma (ref). We previously demonstrated in transgenic mouse models that CD8+ T cells are enriched at pathological lesions and play an essential role in AD progression (cell reports 2023). However, the precise mechanisms governing their recruitment and their interplay with resident central nervous system (CNS) cells remain poorly understood, largely due to profound translational gaps between murine models and human immunity.

The CCL5-CCR5 chemokine signaling axis has emerged as a critical node integrating these cellular responses (13,14). Expressed on both CNS resident cells and peripheral immune cells, the CCR5 receptor and its glial-derived ligand CCL5 facilitate both local neuroinflammation and the chemotactic infiltration of CD8+ T cells (16). Yet, determining the exact, cell-type-specific roles of CCL5-CCR5 and other signaling pathways (20) in human AD has been stalled by a lack of experimental systems capable of faithfully capturing human, multicellular neuroimmune dynamics.

To bridge this fundamental gap, we engineered a modular, human forebrain organoid platform that systematically reconstructs both innate and adaptive neuroimmune microenvironments. Unlike traditional 2D cultures or in vivo murine models, this 3D system structurally and functionally integrates human induced pluripotent stem cell (iPSC)-derived forebrain organoids (21, nat commun 2025) with iPSC-derived microglia (ref) and primary peripheral CD8+ T cells. By exposing this platform to the highly amyloidogenic Aβ42 isoform, we established an experimentally tractable, tri-cellular model that dynamically mimics the human AD neuroinflammatory milieu.

Crucially, the modular nature of this platform permits precise spatiotemporal control over cellular composition, stepwise tracking of T cell infiltration, and real-time functional readouts of neuro-glial-T cell crosstalk. Utilizing this system, we successfully recapitulated key features of human AD immune dysfunction, dissected the human-specific mechanisms translating Aβ pathology into T cell recruitment, and identified and verified novel immunomodulatory cues. Ultimately, this integrated organoid platform provides a comprehensive and versatile framework for dissecting complex neuroimmune interactions and identifying cell-type-specific therapeutic targets in AD.

Result

Aβ42 Impairs Forebrain Organoid Maturation and Induces a Stress-Associated State

To establish a human neural substrate for modeling Alzheimer’s disease (AD), we differentiated forebrain organoids from human iPSCs and exposed them to Aβ42 oligomers for 8 days beginning at day 90 of differentiation (Fig. 1A). Immunofluorescence staining confirmed the presence of cortical progenitor and neuronal populations, including TBR2+, CTIP2+, and SOX2+ cells (Fig. 1B) as well as SATB2+, CTIP2+, and BRN2+ neurons (Fig. 1C), in both control (CTR) and Aβ42-treated organoids. Forebrain regional identity was maintained following Aβ42 exposure, as evidenced by sustained expression of FOXG1, MAP2, SOX2, and PAX6 (Fig. 1D). Astrocyte-associated populations expressing GFAP and S100B were also present in both conditions (Fig. 1E). Immunostaining with the 6E10 antibody confirmed significant Aβ accumulation in treated organoids (Fig. 1F; P < 0.0001).

Aβ impairs forebrain organoid maturation and induces stress-associated states.

(A) Schematic of the experimental design. Forebrain organoids were differentiated and exposed to Aβ42 oligomers for 8 days starting at day 90. (B–C) Immunofluorescence staining of cortical progenitor and neuronal layer markers in control (CTR) and Aβ42-treated organoids. Expression of TBR2, CTIP2, and SOX2 (B), and SATB2, CTIP2, and BRN2 (C), confirms the presence of cortical progenitor and neuronal populations in both CTR and Aβ42-treated organoids. The section sizes do not represent organoid size difference. (D) Representative immunofluorescence images showing expression of forebrain markers FOXG1, MAP2, SOX2, and PAX6 in CTR and Aβ42-treated organoids, indicating maintenance of forebrain regional identity following Aβ42 exposure. (E) Immunostaining for glial markers GFAP and S100B in CTR and Aβ42-treated organoids, demonstrating the presence of astrocyte-associated populations in both conditions. (F) Detection of Aβ using the 6E10 antibody in CTR and Aβ42-treated organoids. Quantification of 6E10 signal intensity normalized to DAPI shows significantly increased Aβ signal in treated organoids (right; violin plot, P < 0.0001, unpaired t test, n = 4–6 organoids per group). (G) UMAP visualization of integrated single-cell RNA-seq data from day 60 CTR and Aβ42-treated organoids, split by condition and colored by annotated cell types. (H) Comparison of relative cell-type proportions between CTR and Aβ42-treated organoids reveals shifts in the abundance of specific transcriptional states following Aβ treatment. (I) Volcano plot showing differentially expressed genes identified by pseudobulk analysis comparing Aβ42-treated organoids to CTR, highlighting induction of stress-associated genes and reduced expression of neuronal maturation–related genes. (J) Gene set enrichment analysis (GSEA) performed on Gene Ontology Biological Process (GO:BP) terms using differentially expressed genes from pseudobulk analysis comparing Aβ42-treated organoids to CTR. Enriched pathways highlight activation of cellular stress–related processes and depletion of neuronal maturation–associated biological processes in Aβ42-treated organoids. (K) Pseudotime trajectory analysis inferred using Monocle3 and visualized on UMAP embeddings for control (CTR; left) and Aβ42-treated (right) organoids, showing differences in the distribution of cells along the inferred neurogenic trajectory between conditions. (L) Quantification of pseudotime distributions in CTR and Aβ42-treated organoids. Aβ42-treated organoids exhibit a significant shift toward lower pseudotime values compared with CTR (Wilcoxon rank-sum test).

To characterize the transcriptional impact of Aβ42 exposure, we performed single-cell RNA sequencing on day 90 CTR and Aβ42-treated organoids. UMAP visualization of integrated data revealed diverse neural cell types, including radial glia, intermediate progenitor cells (IPCs), migrating neurons, maturing neurons, and upper-layer excitatory neurons (Fig. 1G). Comparison of relative cell-type proportions showed shifts following Aβ42 treatment, including an increase in stress-responsive populations and a decrease in maturing neurons (Fig. 1H). Pseudobulk differential expression analysis revealed upregulation of stress-associated genes and downregulation of neuronal maturation-related genes in Aβ42-treated organoids (Fig. 1I). Gene set enrichment analysis (GSEA) confirmed activation of cellular stress pathways and depletion of neuronal maturation programs (Fig. 1J). Pseudotime trajectory analysis using Monocle3 showed that Aβ42-treated organoids exhibited a significant shift toward lower pseudotime values compared with CTR (Fig. 1K-L; Wilcoxon rank-sum test), indicating that Aβ42 exposure arrests cells in an immature, stress-associated state.

Together, these data establish that Aβ42 exposure alone induces transcriptional stress and impairs neuronal maturation in human forebrain organoids, creating a vulnerable neural substrate for subsequent immune interactions.

A 2D-3D Hybrid Organoid System Enables Functional Interrogation of Aβ42-Induced Synaptic and Network Dysfunction

To enable functional analysis of Aβ42-induced pathology, we developed a 2D-3D hybrid organoid system. Day 90 cortical organoids were either maintained as intact 3D structures for bulk RNA-seq analysis or sectioned into 300-µm slices and cultured on Matrigel-coated plates, preserving three-dimensional cytoarchitecture while providing optical access for imaging (Fig. 2B). Bulk RNA-seq of intact 3D organoids confirmed that Aβ42 treatment induced transcriptional changes consistent with our single-cell data, including upregulation of stress-associated genes and downregulation of maturation-related genes (Fig. 2A).

Development and characterization of a 2D–3D hybrid organoid model for analyzing Aβ42-induced pathology.

(A) Heatmap of the top 30 most significantly differentially expressed genes from bulk RNA-seq analysis of day 90 3D Aβ42-treated and CTR organoids. Expression values are shown as scaled normalized expression across individual samples, with hierarchical clustering applied to both genes and samples. (B) Experimental workflow for the study. Day 90 cortical organoids were either maintained as intact 3D organoids for bulk RNA-seq analysis or sectioned into 300-µm slices and cultured on Matrigel-coated plates to generate a 2D–3D hybrid organoid system for synaptic and functional analyses. Hybrid cultures were treated with 5 µM Aβ42 oligomers for 6 days prior to downstream assays. (C) Representative immunofluorescence images of co-stained Syn, MAP2 with VGLUT or GABA to assess synapse ratio. Scale bar: 10 µm. (D–G) Quantitative analysis of synaptic marker density in CTR and Aβ42-treated 2D–3D hybrid organoids. Measurements of excitatory synapses (VGLUT/synaptophysin) and inhibitory synapses (GABA/synaptophysin) reveal increased excitatory and decreased inhibitory synaptic density in Aβ42-treated organoids. (H) Representative calcium imaging traces recorded from CTR and Aβ42-treated 2D–3D hybrid organoids. Images were acquired at 7.67 s per frame. (I) Quantification of calcium imaging data showing increased spike frequency per cell in Aβ42-treated hybrid organoids compared with CTR. (J) GABA response analysis expressed as the post-/pre-treatment spike count ratio, indicating altered GABA responsiveness in Aβ42-treated hybrid organoids. Data are presented as mean ± SEM. For immunofluorescence and functional analyses, n = 4–6 organoids per group. Statistical significance was assessed using unpaired t tests.

We next used the hybrid system to assess synaptic and network-level consequences of Aβ42 exposure. Immunofluorescence staining for synaptic markers revealed a significant shift in excitatory/inhibitory balance: Aβ42-treated organoids exhibited increased density of VGLUT+ excitatory synapses and decreased density of GABA+ inhibitory synapses, normalized to synaptophysin (Fig. 2C-G). This synaptic imbalance was accompanied by functional network alterations. Calcium imaging revealed significantly increased spike frequency per cell in Aβ42-treated hybrid organoids compared with CTR (Fig. 2H-I). Furthermore, GABA response analysis showed an altered post-/pre-treatment spike count ratio in Aβ42-treated organoids (Fig. 2J), suggesting impaired inhibitory signaling.

These results demonstrate that Aβ42 exposure induces both synaptic imbalance and network hyperactivity in human forebrain organoids, establishing functional deficits that recapitulate key features of AD pathophysiology.

Integration of iPSC-Derived Microglia Modulates Aβ42-Associated Transcriptional States and Neural Activity

Having established a Aβ42-exposed neural substrate with defined functional deficits, we next sought to incorporate innate immunity. We generated iPSC-derived microglia (iMGLs) through sequential differentiation of hematopoietic progenitors to immature and then mature microglia (Fig. 3A). Flow cytometry confirmed that approximately 96% of cells were CD11b+, with ∼50% exhibiting a CD45low phenotype consistent with microglial maturation (Fig. 3B). When co-cultured with Aβ42-treated organoids, iMGLs associated with and engulfed Aβ42 oligomers, as visualized by IBA1 and 6E10 co-staining (Fig. 3B). ELISA quantification of culture media showed that iMGL integration significantly reduced Aβ42 levels compared with organoids alone (Fig. 3C; P < 0.0001), confirming functional Aβ clearance.

Integration of iPSC-derived microglia modulates Aβ42-associated transcriptional states and neural activity in forebrain organoids.

(A) Experimental scheme illustrating the microglia-integrated forebrain organoid model. Forebrain organoids were treated with vehicle (0.1% DMSO; CTR) or Aβ42 oligomers (5 μM) for 6 days, followed by a 48-hour co-culture with iPSC-derived microglia (iMGLs) to generate CTR+iMGL and Aβ42+iMGL conditions. The workflow depicts iPSC differentiation through hematopoietic progenitors, immature microglia, and mature microglia stages. (B) Characterization of iPSC-derived microglia. Flow cytometry analysis (left) shows that approximately 96% of cells are CD11b+ and ∼50% exhibit a CD45^low phenotype, consistent with microglial maturation. Representative immunofluorescence images (right) show iMGLs labeled with IBA1 (green) and Aβ42 detected by 6E10 (red), demonstrating association and engulfment of Aβ42 oligomers by iMGLs. Scale bars: 100 μm (overview), 50 μm (high magnification). (C) Quantification of Aβ42 levels in culture medium by ELISA 48 hours after iMGL integration. Aβ42 levels are significantly reduced in the Aβ42+iMGL condition compared with Aβ42-treated forebrain organoids without microglia (P < 0.0001, unpaired t test, n = 3 samples per group). Data are presented as mean ± SEM. (D) UMAP visualization of integrated single-cell RNA-seq data from Aβ42-treated forebrain organoids (Aβ42) and Aβ42-treated forebrain organoids co-cultured with microglia (Aβ42+iMGL), split by condition and colored by annotated cell types. Major neural progenitor and neuronal populations are present in both conditions, with the appearance of a microglial cluster (yellow) in the Aβ42+iMGL group. (E) Comparison of relative cell-type proportions between Aβ42-treated forebrain organoids with (Aβ42+iMGL) and without (Aβ42) microglial co-culture, revealing shifts in the abundance of specific transcriptional states following microglial integration. (F) Pseudotime trajectory analysis inferred using Monocle3 and visualized on UMAP embeddings for Aβ42+iMGL condition, showing the distribution of cells along inferred developmental trajectories. Colors represent pseudotime values from early (purple/blue) to late (yellow/red) developmental states. (G) Quantification of pseudotime distributions comparing Aβ42-treated forebrain organoids with and without microglial co-culture. Aβ42+iMGL organoids exhibit a significant shift toward higher pseudotime values compared with Aβ42 alone (p = 2.17e-16, Wilcoxon rank-sum test). Box plots show median, quartiles, and individual data points. (H) Cell-cell interaction network analysis showing predicted signaling pathways between cell types in Aβ42+iMGL organoids. Node size and edge thickness represent interaction strength. Colors indicate different cell populations including microglia, various neuronal subtypes, and progenitor cells. (I, J) Microglia outgoing (I) and incoming (J) signaling with environmental cells for various signaling pathways. (K) Experimental design for functional analyses using the 2D–3D hybrid forebrain organoid system.Organoid slices were treated with Aβ42 (5 μM) for 6 days, followed by a 48-hour co-culture with iPSC-derived microglia prior to calcium imaging. (L) Representative calcium imaging field showing neuronal calcium activity labeled with Fluo-4 (green) and iPSC-derived microglia labeled with CellTracker CMTPX (red) in Aβ42-treated hybrid organoids. Scale bar: 100 μm. (M) Quantification of neuronal calcium activity (spike counts per cell) showing no significant difference between CTR+iMGL and Aβ42+iMGL conditions in the presence of microglial co-culture. Individual dots represent single cells; horizontal lines indicate mean ± SEM (n = 3–4 organoids per group). (N) Quantification of microglial calcium activity expressed as the area under the curve (AUC) of ΔF/Fo traces, showing significantly increased microglial calcium activity in Aβ42+iMGL conditions compared with CTR+iMGL (P = 0.0346, unpaired t test). Individual dots represent single microglia; horizontal lines indicate mean ± SEM (n = 3–4 organoids per group). 3K is created in BioRender (https://BioRender.com/6o4kfkx).

To understand how microglia reshape the cellular landscape, we performed single-cell RNA sequencing on Aβ42-treated organoids with and without iMGL co-culture (Aβ42+iMGL and Aβ42 conditions). UMAP visualization revealed the appearance of a distinct microglial cluster in the Aβ42+iMGL group, with all major neural populations preserved (Fig. 3D). Comparison of cell-type proportions showed shifts following microglial integration, including changes in progenitor and neuronal subtypes (Fig. 3E). Strikingly, pseudotime analysis revealed that microglial co-culture significantly shifted cells toward higher pseudotime values compared with Aβ42 alone (Fig. 3F-G; P = 2.17e-16), indicating that microglia promote neuronal maturation and partially reverse the Aβ42-induced immature state.

Cell-cell interaction analysis using CellChat identified extensive signaling networks between microglia and neural populations (Fig. 3H). Microglia exhibited both outgoing signals, including neurotrophic (PTN, IGF1), synaptic (NRXN3-NLGN1), and immunomodulatory (LGALS9-HAVCR2) pathways, and incoming signals from neural progenitors and neurons (Fig. 3I-J), revealing bidirectional communication.

We next assessed the functional impact of microglial integration using the 2D-3D hybrid system (Fig. 3K). Calcium imaging with Fluo-4-labeled neurons and CellTracker CMTPX-labeled microglia (Fig. 3L) revealed that, in the presence of microglia, the Aβ42-induced neuronal hyperactivity was normalized: there was no significant difference in spike frequency between CTR+iMGL and Aβ42+iMGL conditions (Fig. 3M). However, microglia themselves exhibited significantly increased calcium activity in the Aβ42+iMGL condition compared with CTR+iMGL (Fig. 3N; P = 0.0346), indicating that while microglia protect neurons from hyperactivity, they become activated in the Aβ environment.

These data demonstrate that microglia integrate into forebrain organoids, clear Aβ, promote neuronal maturation, and normalize network activity, yet simultaneously enter a hyperactive state, revealing their dual functionality in AD pathology.

Microglia Drive CD8+ T Cell Recruitment and Establish a Pathological Neuroimmune Tri-Cellular Unit

The persistent hyperactive state of microglia in Aβ42 environments prompted us to examine their inflammatory output. Conditioned media from CTR+iMGL and Aβ42+iMGL organoids were analyzed for cytokine and chemokine secretion. Aβ42+iMGL organoids exhibited significantly elevated levels of CCL4, CCL5, IL-1β, and TGF-β1, with no change in CXCL10 (Fig. 4A-E). These chemokines are known chemoattractants for adaptive immune cells, particularly CD8+ T cells.

Aβ42-treated forebrain organoids with microglial integration promote inflammatory signaling, CD8+ T cell recruitment, and microglia–T cell functional interactions.

(A–E) Quantification of inflammatory cytokine and chemokine secretion from forebrain organoids co-cultured with microglia. Relative secretion levels of CCL4 (A), CCL5 (B), IL-1β (C), TGF-β1 (D), and CXCL10 (E) were measured in conditioned media from CTR+iMGL and Aβ42+iMGL groups. Data are presented as mean ± SEM. Statistical significance was assessed using unpaired t tests. (F) Experimental schematic of the 2D–3D hybrid forebrain organoid model used to assess CD8+ T cell migration. Forebrain organoids were treated with Aβ42 for 6 days and co-cultured with iPSC-derived microglia (iMGLs), followed by transwell-based migration assays using CD8+ T cells with or without chemokine receptor inhibitors targeting CCR5, CXCR3, and CCR1. (G) Quantification of CD8+ T cell migration in the absence (−MG) or presence (+MG) of microglial co-culture. No significant difference in migration was observed between CTR and Aβ42 conditions without microglia, whereas Aβ42-treated forebrain organoids with iMGLs exhibited significantly increased CD8+ T cell migration. (H) Effects of chemokine receptor inhibition on CD8+ T cell migration toward Aβ42+iMGL forebrain organoids. Blockade of CCR5, CXCR3, or CCR1 on CD8+ T cells reduced migration compared with the untreated Aβ42+iMGL condition. (I) Experimental design for calcium imaging analysis of microglia–T cell interactions. iMGLs were integrated into Aβ42-treated forebrain organoids on day 6, followed by addition of CD8+ T cells on day 7, with calcium imaging performed on day 8. (J) Representative multi-channel imaging showing CD8+ T cells (CellTrace Violet, blue), intracellular calcium signals (Fluo-4 AM, green), iMGLs (CellTracker CMTPX, magenta), and Aβ42 (647 nm). Scale bar: 50 μm. (K) Time-lapse calcium imaging illustrating heterogeneous microglial activation states in the presence of CD8+ T cells. Example regions show iMGLs engaging in direct T cell interactions or Aβ42-associated activity over a 13-minute imaging window. Scale bar: 50 μm. (L) Quantitative analysis of microglial calcium activity comparing Aβ42+iMGL and Aβ42+iMGL+T conditions. Metrics include spike count per cell, area under the curve (AUC), and spike amplitude. Individual dots represent single iMGLs. (M) Quantitative comparison of microglial calcium activity between CTR+iMGL+T and Aβ42+iMGL+T conditions using the same metrics as in (L). (N) Quantitative analysis of CD8+ T cell calcium activity comparing CTR+iMGL+T and Aβ42+iMGL+Tconditions, including spike count per cell, AUC, and spike amplitude. Calcium imaging data were acquired at 3.8-s intervals. (O) Schematic overview of the integrated human forebrain organoid platform. Aβ42-treated forebrain organoids are converted to a 2D–3D hybrid system enabling systematic integration of iPSC-derived microglia and CD8+ T cells. The platform supports multi-modal analysis including single-cell transcriptional profiling, real-time calcium imaging, and assessment of immune-neural cellular interactions. Key features include human-specific cellular components, multi-cellular complexity, systematic component addition, and multiple functional readouts. Data are presented as mean ± SEM. Statistical significance was assessed using unpaired t tests. For cytokine analyses, n = 3 samples per group. For migration and calcium imaging analyses, n = 3–4 organoids per group, with individual dots representing single cells where indicated. 4O is created in BioRender (https://BioRender.com/wgpmokj).

To test whether microglial-derived signals recruit CD8+ T cells, we established a transwell migration assay using the 2D-3D hybrid system (Fig. 4F). In the absence of microglia, CD8+ T cells showed no significant migration toward either CTR or Aβ42-treated organoids (Fig. 4G). However, when microglia were present, Aβ42-treated organoids induced significantly increased T cell migration compared with CTR+iMGL (Fig. 4G). This recruitment was mediated by chemokine receptors on T cells: pharmacological inhibition of CCR5, CXCR3, or CCR1 significantly reduced migration toward Aβ42+iMGL organoids (Fig. 4H).

We next established a tri-culture system to visualize and quantify microglia-T cell interactions (Fig. 4I). Forebrain organoids were treated with Aβ42 for 6 days, co-cultured with iMGLs for 48 hours, and then CD8+ T cells were added 24 hours prior to calcium imaging. Multi-channel imaging enabled simultaneous visualization of T cells (CellTrace Violet), calcium activity (Fluo-4 AM), microglia (CellTracker CMTPX), and Aβ42 (647 nm) (Fig. 4J). Time-lapse imaging revealed heterogeneous microglial activation states, with some microglia engaging in direct T cell interactions and others exhibiting Aβ42-associated activity (Fig. 4K).

Quantitative analysis showed that the presence of T cells further amplified microglial hyperactivity: Aβ42+iMGL+T organoids exhibited significantly increased microglial spike count, area under the curve (AUC), and spike amplitude compared with Aβ42+iMGL without T cells (Figs. 4L, 4M). Microglial activity was also significantly higher in Aβ42+iMGL+T compared with CTR+iMGL+T (Fig. 4N), confirming that the Aβ environment drives sustained microglial activation even in the presence of T cells. CD8+ T cells themselves exhibited calcium activity that was modulated by the environment, with trends toward increased activity in the Aβ42 condition (Fig. 4N).

Together, these data establish a tri-cellular neuroimmune unit in which: (1) Aβ42 creates a stressed neural substrate, (2) microglia respond by clearing Aβ and promoting maturation but become hyperactive and secrete inflammatory chemokines, (3) these chemokines recruit CD8+ T cells via CCR5/CXCR3 signaling, and (4) recruited T cells further amplify microglial activation, creating a positive feedback loop of neuroinflammation (summarized in Fig. 4O).

This modular human forebrain organoid platform, incorporating both innate and adaptive immunity, enables systematic dissection of multicellular neuroimmune interactions in AD and provides an experimentally tractable system for identifying cell-type-specific therapeutic targets.

Discussion

The neuroimmune axis is central to Alzheimer’s disease (AD) pathogenesis, yet the human-specific mechanisms translating amyloid pathology into sustained neuroinflammation remain poorly understood. By establishing a modular human forebrain organoid platform that sequentially integrates innate and adaptive immunity, we successfully deconstructed the tri-cellular interactions underlying AD neuroinflammation. This stepwise approach revealed a multi-stage positive feedback loop: (i) Aβ42 exposure induces neuronal stress; (ii) microglia clear Aβ (Fig. 3B-C) and normalize neuronal hyperactivity (Fig. 3M) but enter a chronically active, chemokine-secreting state (Fig. 3N, 4A-E); (iii) these microglial signals recruit CD8+ T cells via CCR5/CXCR3 pathways (Fig. 4H); and (iv) the recruited T cells subsequently amplify microglial hyperactivity (Fig. 4L). Crucially, T cells failed to migrate to Aβ-treated organoids in the absence of microglia (Fig. 4G). This provides the first human-specific mechanistic demonstration that microglial activation serves as the essential bridge between amyloid pathology and the adaptive immune recruitment recently documented in human AD brains.

Furthermore, our findings experimentally resolve the functional duality of microglia in AD. Within our platform, microglia simultaneously executed protective functions—engulfing Aβ and promoting neuronal maturation (Fig. 3F-G)—while initiating pathological inflammatory cascades. This duality indicates that microglial activation is not a binary state but a spectrum, where sustained Aβ stress drives the transition from a protective to a pathological, T-cell-recruiting state. We identified the CCL5-CCR5 axis as the master integrator of this neuroimmune transition. Microglia-derived CCL5 directly drives CD8+ T cell chemotaxis, while cell-cell interaction analyses revealed extensive CCR5-associated signaling with neurons and progenitors (Fig. 3H-J). Inhibiting CCR5 successfully blocked T cell recruitment (Fig. 4H) but also impacted cell-specific autophagy, indicating that CCR5 acts as a pleiotropic regulator of neuroimmune function rather than a simple chemoattractant. Repurposing clinical CCR5 antagonists (e.g., Maraviroc) or CXCR3 inhibitors offers an immediate therapeutic avenue to interrupt this pathological cascade; however, our data caution that precise modulation is necessary to suppress inflammatory outputs without sacrificing the protective clearance and maturation functions of microglia. While our 2D-3D hybrid platform captures human-specific cellular features, transcriptomic signatures, and real-time multicellular interactions unmatched by rodent or conventional in vitro models, we acknowledge several limitations. The current system lacks vasculature and additional peripheral immune populations, utilizes a single iPSC line, and reflects an acute timeline rather than decades-long disease progression. Future iterations incorporating multi-donor genetic variability - particularly AD risk variants - and extended culture periods will be critical to fully modeling disease evolution.

The modular, tri-cellular platform described here has implications beyond AD. The same approach could be applied to model other neurodegenerative and neuroinflammatory conditions, including Parkinson’s disease, amyotrophic lateral sclerosis, multiple sclerosis, and neuropsychiatric disorders with immune components. By enabling systematic addition of specific immune populations - for example, different T cell subsets, B cells, or peripheral monocytes - this platform can be adapted to address diverse questions about neuroimmune interactions in health and disease.

From a therapeutic perspective, our findings identify multiple intervention points. CCR5 and CXCR3 antagonists, already in clinical use or development for other indications, could be tested for their ability to block T cell infiltration and interrupt the neuroinflammatory feedback loop. Strategies to preserve microglial protective functions (Aβ clearance, neuronal maturation) while suppressing their pathological outputs (chemokine secretion, T cell recruitment) represent a particularly attractive therapeutic goal. The platform’s compatibility with high-content imaging and multi-well formats positions it for medium-throughput drug screening, and its human relevance may improve translation compared with traditional animal models.

Despite these constraints, the identification of a tri-cellular positive feedback loop offers a mechanistic explanation for the persistence of neuroinflammation in AD, even after initial pathological triggers may have subsided. Beyond AD, this modular, high-content-compatible platform provides a versatile framework to systematically dissect neuroimmune complexity, enabling the stepwise addition of distinct immune subsets to accelerate targeted drug discovery across a broad spectrum of neurodegenerative and neuropsychiatric disorders.

Methods

Animals

Male 5×FAD transgenic mice (Tg6799 line) expressing five familial AD mutations (APP K670N/M671L + I716V + V717I and PS1 M146L + L286V) and age-matched wild-type littermates were used at 8 months of age and bought from GUANGZHOU JINWEI BIOSCIENCE. Animals were housed under controlled conditions (22°C, 12-hour light/dark cycle) with ad libitum access to standard laboratory chow and water. All experimental procedures were performed in compliance with protocols approved by the Animal Experimentation Ethics Committee at Tsinghua Shenzhen International Graduate School.

Cerebral organoids, cell lines, medium and reagent

The protocol for the establishment of cerebral organoids was adapted from methods described in our previous work21. The cell lines used in this study included the H9 line, obtained from WiCell (Catalog # WA09), the iPSC cell line DYR0100, acquired from the National Collection of Authenticated Cell Cultures (Catalog # SCSP-1301) and DXR0109B, acquired from ATCC (Catalog #ACS-1023). Material transfer agreements (MTAs) for both the H9 and iPSC cell lines have been duly executed. The media and reagents employed for culturing these cell lines and for cerebral organoid development are detailed in our prior work.

iPSC-derived microglia

Both induced and embryonic microglia were generated following a two-step induction process. First, hematopoietic cells were derived using the STEMdiff™ Hematopoietic Kit (STEMCELL Technologies, Catalog # 05310_C). These cells were then differentiated into microglia using the STEMdiff™ Microglia Differentiation Kit (Catalog # 100-0019) and matured using the STEMdiff™ Microglia Maturation Kit (Catalog # 100-0020_C). All procedures were performed in strict accordance with the manufacturer’s protocols to ensure proper induction and maturation of microglia.

2D/3D neural-immune model establishment in Aβ42 context

Cerebral organoids aged 90 days were sliced into 300 μm sections using a vibratome, following the methodology described in our previous work. The sliced COs were then cultured on 1% Matrigel-coated 12-well confocal plates, with one slice per well, to facilitate immunofluorescence imaging. Within 24 hours of culturing, neurons and astrocytes migrated outwards from the center of the COs. The culture was supplemented with 5 μM

β-amyloid (1-42) (MedChemExpress, Catalog # HY-P1363), and the medium was changed every two days. After 5 days, 1×105 induced microglia (iMGLs) were added to each well to establish the neural immune model. The samples were fixed for further analysis 48 hours after the introduction of iMGLs.

Isolation of human PBMC and CD8+ T cells

Blood specimens were collected from healthy donors at Zhangzhou People’s Hospital. Peripheral blood was collected from individuals over 18 years of age in 10-ml tubes containing 5 mM EDTA. The samples were transported on ice and processed within 1-3 hours of collection. PBMCs were isolated using density gradient centrifugation with Lymphoprep™ density gradient medium (1.077 g/mL). The PBMCs were then handled with care to maintain their viability and integrity.

The human CD8+ T cells were isolated from the PBMCs using the EasySep™ Direct Human CD8+ T Cell Isolation Kit (STEMCELL Technologies, Catalog # 19663) following the manufacturer’s protocols to ensure efficient and consistent isolation of CD8+ T cells.

Flow cytometry

Validation of isolated and migrated CD8+ T cells was conducted using APC/Cyanine7 anti-human CD8 Antibody (Biolegend, Catalog # 344714). For the verification of mature microglia, FITC anti-human CD45 Antibody (Biolegend, Catalog # 304006) and PE anti-human CD11b Antibody (Biolegend, Catalog # 301306) were utilized. The compensation matrix was established and positive peak gates were set based on the respective isotype controls. The isotype controls used were APC/Cyanine7 Mouse IgG1, κ Isotype Ctrl Antibody, Catalog # 400127; PE Mouse IgG1, κ Isotype Ctrl Antibody, Catalog # 400111; FITC Mouse IgG1, κ Isotype Ctrl Antibody, Catalog # 400107.

The cells were incubated with the specified antibodies for 40 minutes on ice to ensure adequate labeling. Before analysis, the cells underwent filtration through a 300-mesh Falcon centrifuge tube to eliminate aggregates and ensure a single-cell suspension. All flow cytometry data were then analyzed using FlowJo™ Software v10.10. for detailed phenotypic characterization.

Cell Tracker staining

Before conducting calcium imaging, Cell Tracker dyes were used to stain CD8+ T cells and iMGLs for better identification. CD8+ T cells were stained with CellTrace™ Violet (dilution 1:1000, Invitrogen, Catalog # C34557), and iMGLs were labeled with CellTracker™ CMTPX (dilution 1:1000, Invitrogen, Catalog # C34552). The cells were then incubated in the staining solution at 37 °C for 30 minutes. After incubation, the staining solution was removed, and the cells were cultured for future use within the next week.

Human inflammatory cytokine array

For each experimental group, 8 ml of medium was collected and concentrated using Ultrafiltration Centrifugal Tubes (Millipore, Catalog # UFC903096). The medium was reduced to a volume of 200 μl through ultrafiltration. This concentrated sample was then used to perform the Cytokine Array analysis using the Human Cytokine Antibody Array (Membrane, 42 Targets) kit (Abcam, Catalog # ab133997), strictly following the manufacturer’s protocols.

Raw images of the cytokine arrays have been made available in the source data, which can be accessed via the provided Github link: https://github.com/FengYilinElaine/2-3D-AD-oranoids.

Immunohistochemistry

Validation of COs was based on previous work. The primary antibodies used in this study are listed in Extended Table 1.

Quantitative analysis of Aβ42 levels

Quantitative analysis of Aβ42 levels was performed using the Amyloid beta 42 Human ELISA Kit (Invitrogen, Catalog # KHB3441).

Bulk RNA sequencing analysis

Bulk RNA sequencing analysis was performed on 3D cerebral organoids, iMGLs and eMGLs, as described in the results section. The samples were collected quickly and flash-frozen in liquid nitrogen to preserve the RNA integrity. After freezing, the samples were transported on dry ice to BGI for RNA extraction and sequencing. All RNA analyses, including data processing and interpretation, were conducted on the Dr. Tom platform (BGI).

Single-cell RNA-seq Library Preparation and Sequencing

For isolation of iPSC induced forebrain cells, they were collected from CTR organoids, CHOL organoids, and semaglutide-treated CHOL+SEMA organoids at day 30. Two biological replicates were included in each group. Organoids were pooled respectively and washed three times with pre-cold PBS. Single-cell suspensions were prepared by incubating with 0.5 mg/mL digestive enzymes 1 for 30 min at 37 °C in a water bath. After, suspensions were filtered through a 70 μM cell strainer (BD Falcon, 352350). After centrifugation at 500 g for 5 min, the cell precipitated was resuspension with PBS+0.01%BSA.

The MGI DNBelab C-TaiM4 platform utilizes microfluidic technology to encapsulate single cells with mRNA capture magnetic beads containing unique cell barcodes (Cell Barcodes) and unique molecular identifiers (UMIs) within oil droplets. The platform employs a dual-magnetic bead system (mRNA capture beads and droplet-identification beads) to enhance capture efficiency and accuracy. Then, cell lysis was completed in the droplet, and mRNA molecules were captured by Oligo (dT) on the magnetic beads and reverse transcribed to generate cDNA with cell barcodes and UMIs.

For library construction, the cDNA was purified, amplified, and ligated with sequencing adapters to generate the final sequencing library. Library quality was checked using an Agilent 2100 Bioanalyzer (Agilent Technologies, USA). Finally, qualified libraries were sequenced in PE100 mode using DNBSEQ-T7 from BGI. The cDNA library sequencing depth was approximately 450M Reads/library, and the Oligo library sequencing depth was approximately 50M Reads/library. scRNA-seq were performed by Hangzhou Astrocyte Technology Co,.Ltd.

scRNA-seq Data Pre-processing and Quality Control

Using the official BGI software DNBC4Tools with parameter settings, we performed filtering, alignment, quantification, and cell identification/recovery of raw data. After transcriptomic normalization processing, gene expression matrix files for each cell were obtained. The resulting gene expression matrices were imported into R (v.4.4.1) and processed using the Seurat package (v.5.3.1) for integration and quality control. rigorous filtering criteria were applied to retain high-quality cells: (1) 200 < total detected genes < 12,500; (2) total unique molecular identifier (UMI) counts > 500; (3) log10(GenesPerUMI) > 0.8; and (4) mitochondrial mRNA content < 20% and ribosomal mRNA content < 30%.

Data analysis: Colocalization ratio of images

Confocal images were initially processed by selecting regions of interest (ROIs) and applying adaptive thresholding for binarization. Different adaptive threshold methods were used for various antibodies. The binarized images were then batch-processed using the ‘Colocalization Finder’ plugin in Fiji. A script for processing colocalization in Fiji, along with detailed methods, is available at the provided GitHub link.

Data analysis: Excitatory synapse and inhibitory synapse

The images underwent preliminary processing, which involved selecting regions of interest (ROIs) and applying adaptive thresholding techniques. Synapse numbers were quantified using the ‘Puncta Analyzer’ plugin in Fiji.

Calcium imaging and analysis

Brain organoids were cultured using the microwell air-liquid interface attachment method 43. Perform two medium changes using BrainPhys™ Neuronal Medium before beginning incubation with Flou-4 calcium indicator (Invitrogen, Catalog # F14201). Using the complete culture medium, prepare a 5 µM Flou-4 solution, add 200 µL to a single well and incubate for 2 hours in the incubator. Make two medium changes and add 500 μL of BrainPhys™ Neuronal Medium (STEMCELL Technologies, Catalog # 05790) 30 minutes prior to baseline imaging. Baseline imaging was performed using a confocal microscope (NikonA1R) for 5 minutes with no delay between frames, and the frame rate was 7.67. After baseline imaging, neurotransmitter drugs, GABA (Sigma-Aldrich, Catalog # 56-12-2) 100 μM, were added for final concentration. After 1 minute incubation, change the medium twice and add 200 μL BrainPhys™ Neuronal Medium. Immediately image, keeping the field of view and parameters the same as the baseline image.

For batch processing in calcium imaging data analysis, CNMF-E44 was used. The code for parameters (getmat.m) can be found in the source. The sum of neuron.S, representing all neurons in the view, is recorded as the total number of spikes after processing. The number of active cells is recorded as the number of rows of neuron.S, i.e. the ROI where the spike is detected. The calcium trace overlay plot is presented by CNMFeClustering29. Calcium imaging analysis of microglia was initially conducted by isolating each microglial cell based on the maximum intensity projection of image sequences in the Cell Tracker channel. Subsequently, the Fluo-4 AM sequence image data for each isolated microglia was analyzed. Spikes were defined and quantified using a specific code, which is provided for reference.

Statistical analysis

All statistical analyses in this study were conducted using commercially available software, namely Prism and Excel. Data from replicates are presented as mean ± standard error of the mean (SEM). The methodology for analyzing differences between multiple groups is detailed in the figure legends accompanying each graph. For the creation of graphical representations, Adobe Illustrator 2024 and Biorender were utilized.

eLife Assessment

This study presents a valuable human organoid platform for investigating neuron-immune interactions in Alzheimer's disease and modeling the interplay between innate and adaptive immunity in the context of amyloid pathology. The system has the potential to advance the field by enabling the study of human-specific neuroimmune interactions that are difficult to recapitulate in rodent models; however, the evidence supporting several central conclusions remains incomplete. Key claims, including the causal role of microglia-derived chemokines in T cell recruitment and the existence of a neuroinflammatory feedback loop, rely primarily on correlative observations and lack the mechanistic experiments necessary to establish causality; in addition, a major confounding factor, the non-autologous nature of the cellular components, is not adequately addressed. Despite these limitations, the study will be of considerable interest to researchers in neuroimmunology and Alzheimer's disease.

Reviewer #1 (Public review):

Summary:

A growing body of evidence indicates that Alzheimer's disease is not simply a disease of neurons accumulating toxic protein aggregates, but one in which the immune system, both its resident brain component and its circulating peripheral arm, plays an active and sustained role. Understanding how these two immune compartments interact with one another and with diseased neural tissue has been hampered by the fact that the mouse immune system differs fundamentally from the human one in ways likely to matter for disease progression. The authors set out to address this gap by building a modular laboratory model that brings together three human cell types in a three-dimensional setting: brain organoids derived from human stem cells to provide a neural substrate, stem cell-derived brain immune cells (microglia) to represent the resident immune compartment, and circulating immune cells (CD8-positive T cells) harvested from human blood to represent the peripheral adaptive immune response. By exposing this tri-cellular system to a toxic form of amyloid protein, the hallmark aggregating molecule of Alzheimer's disease, the authors aimed to dissect, step by step, how microglia respond to amyloid stress, what inflammatory signals they release as a consequence, and whether those signals are sufficient to attract T cells into the neural environment. They further aimed to test whether blocking the molecular receptors that guide T cell movement could interrupt this process, with the broader goal of positioning the platform as a tool for human-relevant drug screening.

Strengths

The conceptual architecture of the platform is one of its clearest strengths. The decision to add immune components in a stepwise, modular fashion, first characterising the neural response to amyloid, then adding microglia, then adding T cells, makes it possible to attribute observed changes to specific cellular contributions in a way that a more complex all-at-once model would not allow. This staged design is well thought-through, and its logic is clearly communicated. The combination of single-cell transcriptional profiling, calcium imaging for real-time functional readouts, transwell migration assays, and protein secretion measurements gives the study a genuinely multi-modal character that goes beyond what purely transcriptomic or purely imaging-based approaches can offer. The observation that T cells failed to migrate toward amyloid-treated organoids in the absence of microglia is a clean and conceptually important result, clearly supporting the idea that the resident immune response acts as an intermediary between amyloid pathology and the recruitment of peripheral immune cells. The identification of specific chemokine receptor pathways mediating T cell movement and the demonstration that pharmacological blockade of those receptors reduces migration and provide a degree of mechanistic resolution useful for thinking about future therapeutic strategies.

Weaknesses

Despite these strengths, several aspects of the work as presented substantially limit the confidence one can place in its conclusions.

The most consequential issue concerns the origin of the cells used in the model. The three cellular components: the brain organoids, the microglia, and the T cells are derived from genetically unrelated individuals. The T cells, in particular, come from healthy blood donors unrelated to the stem cell lines used to generate the neural tissue. This means the immune cells and the tissue they are interacting with carry different molecular identity markers (the proteins that the immune system uses to distinguish self from non-self). In this setting, any T cell activation or directed movement could reflect a generic rejection-like response to foreign tissue rather than a disease-relevant, chemokine-directed recruitment process. This is not a subtle concern: it represents a fundamental ambiguity at the heart of the model's central finding, and it is not acknowledged anywhere in the manuscript. For the transwell migration data to be interpretable as a model of Alzheimer's disease rather than of immune incompatibility, the authors would need to demonstrate that migration is driven by the specific chemokine environment and not by the genetic mismatch between cells, for example, using cells from the same donor or from matched donors, or by showing that blocking identity-marker recognition does not alter migration.

A related concern is that the T cells used are from healthy individuals, whereas T cells from people with Alzheimer's disease are known to differ in their activation state, surface receptor expression, and functional behaviour. The platform cannot yet claim to model the specific T cell biology of Alzheimer's disease until disease-relevant T cells are incorporated.

Beyond this foundational issue, the study frequently describes findings in causal terms that the experimental design does not support. The resident immune cells are said to "drive" T cell recruitment and "establish" a feedback loop. These are strong mechanistic claims. The evidence presented indicates that when microglia are present, more T cells migrate, and that blocking T cells receptors reduces migration. What is missing is direct evidence that the specific molecules measured, particularly the chemokines CCL4 and CCL5, are the agents responsible, as opposed to other signals also present in the conditioned environment. No experiment directly neutralises these chemokines to test whether their removal is sufficient to abolish T cell recruitment. Without such an experiment, the receptor-blocking data show only that the receptors matter, not that the measured ligands are the ones activating those receptors.

The abstract describes one particular molecule, CXCL10, as a contributor to T cell recruitment, but the data in the paper itself show no significant change in CXCL10 levels between conditions. This discrepancy between the abstract and the results is misleading to readers who may not read the figures in detail.

The single-cell sequencing data, which form the basis for claims about changes in cell populations following amyloid treatment or microglia addition, are presented without validation of the cell type labels against established reference datasets from human brain tissue. The proportional shifts in cell populations between conditions (Figures 1H and 3E) are described as significant findings but are shown without any statistical test appropriate for this type of compositional data. Comparisons of cell-type proportions derived from single-cell sequencing require specialised statistical approaches that account for the interdependence of proportions and the variability between samples; standard tests are not appropriate here, and none are applied.

There is also an unresolved inconsistency in the age at which the organoids were analysed by single-cell sequencing: the text states day 90, while the figure legend states day 60, and the methods section contains a passage describing experimental conditions (including a cholesterol treatment and a drug called semaglutide) that are entirely unrelated to this study and appear to have been copied from a different manuscript. These issues raise concerns about the rigour of the manuscript preparation and should be corrected.

Finally, the sample sizes underpinning several key conclusions are small (typically three to four organoids per group), particularly for the protein-secretion measurements used to identify the inflammatory signals responsible for T cell recruitment. While organoid studies are inherently limited in scale, the strength of the mechanistic claims made here would benefit from larger sample size or independent experimental replication.

Conclusion:

The authors have built a platform that is conceptually well-conceived and generates data consistent with a role for microglia in bridging amyloid pathology and T cell recruitment. In that sense, they have made meaningful progress toward their stated aims. However, the platform, as described, cannot yet deliver the human-specific mechanistic insight it claims to provide, primarily because the non-autologous configuration of the model introduces an uncontrolled variable that confounds the interpretation of the immune interaction data. The claim to have provided "the first human-specific mechanistic demonstration" of microglial activation as a bridge between amyloid pathology and adaptive immune recruitment is not supported by the evidence presented. The data are consistent with this interpretation but do not establish it.

The general approach, building increasingly complex human neural-immune models by adding components in a controlled, stepwise manner, is a valuable direction for the field and one that other groups working on neuroinflammation will find useful to consider. The combination of live calcium imaging and transcriptional profiling in the same experimental system is a practical contribution that demonstrates the kind of multi-modal readout this class of model can support. If the autologous confound is resolved in future iterations and if the mechanistic claims are grounded in more direct experimental evidence, this type of platform could become a genuinely useful tool for investigating human neuroimmune biology and for screening candidate therapeutic compounds in a human-relevant context. As currently presented, however, readers and researchers considering adopting this approach should be aware that the immune interaction data may reflect genetic mismatches between cell sources rather than disease-specific biology, and that the causal conclusions drawn from the chemokine and migration data go beyond what the experiments can support.

Reviewer #2 (Public review):

Summary:

In this study, the authors developed a human forebrain organoid model incorporating both iPSC-derived microglia and CD8⁺ T cells, enabling them to recreate and investigate multicellular aspects of AD pathology in a human-relevant system.

Their findings show that microglia help clear amyloid-β deposits, but they also promote inflammatory responses. Activated microglia recruit CD8⁺ T cells by releasing the chemokines CCL4, CCL5, and CXCL10, which signal through the receptors CCR1/CCR5 and CXCR3. Pharmacological inhibition of CCR5 or CXCR3 prevents T-cell recruitment and alters autophagy pathways in a microglia-dependent manner.

Strengths:

The study presents a versatile human organoid platform for investigating neuron-immune interactions in Alzheimer's disease. It highlights the critical role of microglia-driven recruitment of CD8⁺ T cells in sustaining neuroinflammation and identifies CCR5 and CXCR3 signaling pathways as promising therapeutic targets for neuroinflammatory conditions.

This study is interesting and presents novel findings supported by state-of-the-art approaches, including single-cell RNA sequencing, a three-dimensional cerebral organoid model, and co-culture systems involving two distinct immune cell populations.

Weaknesses:

Several aspects of the study require clarification and further improvement. For example:

(1) Figure 1H is missing statistical analyses.

(2) The scRNA-seq analysis shows a reduction in the proportion of cells occupying transcriptional states associated with later pseudotime values, which the authors interpret as evidence that Aβ treatment inhibits neuronal maturation. However, the data presented do not appear sufficient to support this conclusion. An alternative explanation is that Aβ preferentially affects the survival of more mature neuronal populations, leading to their depletion, consequently, an apparent enrichment of cells at earlier pseudotime states. Therefore, the observed pseudotime shift does not necessarily demonstrate impaired maturation per se. The authors should revise the interpretation of these results in the first paragraph and either provide additional evidence supporting a maturation defect or discuss alternative explanations such as selective loss of mature neurons.

(3) A similar concern applies to the scRNA-seq data presented in Figure 3. The authors interpret the shift toward later pseudotime states in the presence of microglia as evidence of enhanced neuronal maturation. However, the data do not exclude alternative explanations. For instance, microglia may preferentially promote the survival of more mature neuronal populations or protect them from cell death, thereby increasing their relative abundance in the dataset. Consequently, the observed pseudotime distribution cannot be taken as direct evidence of enhanced maturation. The authors should revise their interpretation accordingly and discuss the possibility that the observed effect reflects differential survival rather than accelerated neuronal maturation.

(4) In Figures 4A-E, the authors should report the levels of the secreted proteins in pg/mL instead of relative values, as this would better reflect the actual amounts produced. In Figure 4H, the inhibitor-treated control T-cell samples should be included. Furthermore, it should be explicitly stated that the inhibitor-treated data points currently shown refer to T cells cultured in the presence of myeloid Aβ.

Author response:

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer #1 (Public review):

Summary:

A growing body of evidence indicates that Alzheimer's disease is not simply a disease of neurons accumulating toxic protein aggregates, but one in which the immune system, both its resident brain component and its circulating peripheral arm, plays an active and sustained role. Understanding how these two immune compartments interact with one another and with diseased neural tissue has been hampered by the fact that the mouse immune system differs fundamentally from the human one in ways likely to matter for disease progression. The authors set out to address this gap by building a modular laboratory model that brings together three human cell types in a three-dimensional setting: brain organoids derived from human stem cells to provide a neural substrate, stem cell-derived brain immune cells (microglia) to represent the resident immune compartment, and circulating immune cells (CD8-positive T cells) harvested from human blood to represent the peripheral adaptive immune response. By exposing this tri-cellular system to a toxic form of amyloid protein, the hallmark aggregating molecule of Alzheimer's disease, the authors aimed to dissect, step by step, how microglia respond to amyloid stress, what inflammatory signals they release as a consequence, and whether those signals are sufficient to attract T cells into the neural environment. They further aimed to test whether blocking the molecular receptors that guide T cell movement could interrupt this process, with the broader goal of positioning the platform as a tool for human-relevant drug screening.

Strengths

The conceptual architecture of the platform is one of its clearest strengths. The decision to add immune components in a stepwise, modular fashion, first characterising the neural response to amyloid, then adding microglia, then adding T cells, makes it possible to attribute observed changes to specific cellular contributions in a way that a more complex all-at-once model would not allow. This staged design is well thought-through, and its logic is clearly communicated. The combination of single-cell transcriptional profiling, calcium imaging for real-time functional readouts, transwell migration assays, and protein secretion measurements gives the study a genuinely multi-modal character that goes beyond what purely transcriptomic or purely imaging-based approaches can offer. The observation that T cells failed to migrate toward amyloid-treated organoids in the absence of microglia is a clean and conceptually important result, clearly supporting the idea that the resident immune response acts as an intermediary between amyloid pathology and the recruitment of peripheral immune cells. The identification of specific chemokine receptor pathways mediating T cell movement and the demonstration that pharmacological blockade of those receptors reduces migration and provide a degree of mechanistic resolution useful for thinking about future therapeutic strategies.

Weaknesses

Despite these strengths, several aspects of the work as presented substantially limit the confidence one can place in its conclusions.

The most consequential issue concerns the origin of the cells used in the model. The three cellular components: the brain organoids, the microglia, and the T cells are derived from genetically unrelated individuals. The T cells, in particular, come from healthy blood donors unrelated to the stem cell lines used to generate the neural tissue. This means the immune cells and the tissue they are interacting with carry different molecular identity markers (the proteins that the immune system uses to distinguish self from non-self). In this setting, any T cell activation or directed movement could reflect a generic rejection-like response to foreign tissue rather than a disease-relevant, chemokine-directed recruitment process. This is not a subtle concern: it represents a fundamental ambiguity at the heart of the model's central finding, and it is not acknowledged anywhere in the manuscript. For the transwell migration data to be interpretable as a model of Alzheimer's disease rather than of immune incompatibility, the authors would need to demonstrate that migration is driven by the specific chemokine environment and not by the genetic mismatch between cells, for example, using cells from the same donor or from matched donors, or by showing that blocking identity-marker recognition does not alter migration.

A related concern is that the T cells used are from healthy individuals, whereas T cells from people with Alzheimer's disease are known to differ in their activation state, surface receptor expression, and functional behaviour. The platform cannot yet claim to model the specific T cell biology of Alzheimer's disease until disease-relevant T cells are incorporated.

Beyond this foundational issue, the study frequently describes findings in causal terms that the experimental design does not support. The resident immune cells are said to "drive" T cell recruitment and "establish" a feedback loop. These are strong mechanistic claims. The evidence presented indicates that when microglia are present, more T cells migrate, and that blocking T cells receptors reduces migration. What is missing is direct evidence that the specific molecules measured, particularly the chemokines CCL4 and CCL5, are the agents responsible, as opposed to other signals also present in the conditioned environment. No experiment directly neutralises these chemokines to test whether their removal is sufficient to abolish T cell recruitment. Without such an experiment, the receptor-blocking data show only that the receptors matter, not that the measured ligands are the ones activating those receptors.

The abstract describes one particular molecule, CXCL10, as a contributor to T cell recruitment, but the data in the paper itself show no significant change in CXCL10 levels between conditions. This discrepancy between the abstract and the results is misleading to readers who may not read the figures in detail.

The single-cell sequencing data, which form the basis for claims about changes in cell populations following amyloid treatment or microglia addition, are presented without validation of the cell type labels against established reference datasets from human brain tissue. The proportional shifts in cell populations between conditions (Figures 1H and 3E) are described as significant findings but are shown without any statistical test appropriate for this type of compositional data. Comparisons of cell-type proportions derived from single-cell sequencing require specialised statistical approaches that account for the interdependence of proportions and the variability between samples; standard tests are not appropriate here, and none are applied.

There is also an unresolved inconsistency in the age at which the organoids were analysed by single-cell sequencing: the text states day 90, while the figure legend states day 60, and the methods section contains a passage describing experimental conditions (including a cholesterol treatment and a drug called semaglutide) that are entirely unrelated to this study and appear to have been copied from a different manuscript. These issues raise concerns about the rigour of the manuscript preparation and should be corrected.

Finally, the sample sizes underpinning several key conclusions are small (typically three to four organoids per group), particularly for the protein-secretion measurements used to identify the inflammatory signals responsible for T cell recruitment. While organoid studies are inherently limited in scale, the strength of the mechanistic claims made here would benefit from larger sample size or independent experimental replication.

Conclusion:

The authors have built a platform that is conceptually well-conceived and generates data consistent with a role for microglia in bridging amyloid pathology and T cell recruitment. In that sense, they have made meaningful progress toward their stated aims. However, the platform, as described, cannot yet deliver the human-specific mechanistic insight it claims to provide, primarily because the non-autologous configuration of the model introduces an uncontrolled variable that confounds the interpretation of the immune interaction data. The claim to have provided "the first human-specific mechanistic demonstration" of microglial activation as a bridge between amyloid pathology and adaptive immune recruitment is not supported by the evidence presented. The data are consistent with this interpretation but do not establish it.

The general approach, building increasingly complex human neural-immune models by adding components in a controlled, stepwise manner, is a valuable direction for the field and one that other groups working on neuroinflammation will find useful to consider. The combination of live calcium imaging and transcriptional profiling in the same experimental system is a practical contribution that demonstrates the kind of multi-modal readout this class of model can support. If the autologous confound is resolved in future iterations and if the mechanistic claims are grounded in more direct experimental evidence, this type of platform could become a genuinely useful tool for investigating human neuroimmune biology and for screening candidate therapeutic compounds in a human-relevant context. As currently presented, however, readers and researchers considering adopting this approach should be aware that the immune interaction data may reflect genetic mismatches between cell sources rather than disease-specific biology, and that the causal conclusions drawn from the chemokine and migration data go beyond what the experiments can support.

Reviewer #2 (Public review):

Summary:

In this study, the authors developed a human forebrain organoid model incorporating both iPSC-derived microglia and CD8+ T cells, enabling them to recreate and investigate multicellular aspects of AD pathology in a human-relevant system.

Their findings show that microglia help clear amyloid-β deposits, but they also promote inflammatory responses. Activated microglia recruit CD8+ T cells by releasing the chemokines CCL4, CCL5, and CXCL10, which signal through the receptors CCR1/CCR5 and CXCR3. Pharmacological inhibition of CCR5 or CXCR3 prevents T-cell recruitment and alters autophagy pathways in a microglia-dependent manner.

Strengths:

The study presents a versatile human organoid platform for investigating neuron-immune interactions in Alzheimer's disease. It highlights the critical role of microglia-driven recruitment of CD8+ T cells in sustaining neuroinflammation and identifies CCR5 and CXCR3 signaling pathways as promising therapeutic targets for neuroinflammatory conditions.

This study is interesting and presents novel findings supported by state-of-the-art approaches, including single-cell RNA sequencing, a three-dimensional cerebral organoid model, and co-culture systems involving two distinct immune cell populations.

Weaknesses:

Several aspects of the study require clarification and further improvement. For example:

(1) Figure 1H is missing statistical analyses.

(2) The scRNA-seq analysis shows a reduction in the proportion of cells occupying transcriptional states associated with later pseudotime values, which the authors interpret as evidence that Aβ treatment inhibits neuronal maturation. However, the data presented do not appear sufficient to support this conclusion. An alternative explanation is that Aβ preferentially affects the survival of more mature neuronal populations, leading to their depletion, consequently, an apparent enrichment of cells at earlier pseudotime states. Therefore, the observed pseudotime shift does not necessarily demonstrate impaired maturation per se. The authors should revise the interpretation of these results in the first paragraph and either provide additional evidence supporting a maturation defect or discuss alternative explanations such as selective loss of mature neurons.

(3) A similar concern applies to the scRNA-seq data presented in Figure 3. The authors interpret the shift toward later pseudotime states in the presence of microglia as evidence of enhanced neuronal maturation. However, the data do not exclude alternative explanations. For instance, microglia may preferentially promote the survival of more mature neuronal populations or protect them from cell death, thereby increasing their relative abundance in the dataset. Consequently, the observed pseudotime distribution cannot be taken as direct evidence of enhanced maturation. The authors should revise their interpretation accordingly and discuss the possibility that the observed effect reflects differential survival rather than accelerated neuronal maturation.

(4) In Figures 4A-E, the authors should report the levels of the secreted proteins in pg/mL instead of relative values, as this would better reflect the actual amounts produced. In Figure 4H, the inhibitor-treated control T-cell samples should be included. Furthermore, it should be explicitly stated that the inhibitor-treated data points currently shown refer to T cells cultured in the presence of myeloid Aβ.

Recommendations for the authors:

Reviewer #1 (Recommendations for the authors):

(1) Addressing the allogeneic confound

To determine whether observed T cell migration reflects chemokine-driven recruitment rather than allogeneic recognition, the authors should perform one or more of the following: (a) repeat transwell migration assays using HLA-matched donors for T cells and iPSC lines; (b) include a blocking condition using anti-HLA class I antibody to suppress allogeneic recognition and test whether migration is reduced; (c) compare migration toward conditioned media alone (without cells) versus co-culture conditions to establish whether secreted factors are sufficient to drive migration independently of direct cell contact. If HLA-matched material is not currently available, the authors could, at a minimum, use conditioned media transfer experiments to dissociate chemokine-mediated from contact-mediated effects.

We agree this is an important limitation. We have not performed HLA-matched migration assays, anti-HLA class I blockade, or conditioned-media-only migration experiments. We note that in our transwell assay design (Fig. 4G–H), T cells are seeded in the upper chamber and organoids in the lower chamber, separated by a porous membrane; while this precludes direct T cell–organoid cell contact, live organoid/microglial cells are present throughout the assay rather than their conditioned medium alone, so we have not directly tested part (c) as the reviewer describes it.

We would also note that the CD8+ T cells used in this assay were derived from a single healthy donor, while the organoids and iMGLs were derived from a separate iPSC line (DYR0100); the same T cell donor and the same organoid/microglial line were used across all conditions (CTR, Aβ42, Aβ42+iMGL). The T cell–organoid pairing is therefore allogeneic, but identical in every condition. Any contribution of allogeneic recognition to baseline migration should, in principle, be present equally across conditions and would not by itself account for the increased migration specifically toward the Aβ42 and Aβ42+iMGL conditions relative to CTR.

(2) Direct validation of CCL5 as the responsible ligand

To substantiate the claim that microglia-derived CCL5 is responsible for T cell recruitment, a CCL5-neutralising antibody should be added to the transwell assay. Similarly, a CCL4-neutralising antibody should be tested independently. This would allow the authors to attribute the migration effect to specific molecules rather than to the overall conditioned environment and would substantially strengthen the mechanistic interpretation.

We appreciate this comment and agree that ligand-neutralization experiments would substantially strengthen the mechanistic interpretation. We have not performed these experiments in the current study but will consider including CCL5- and CCL4-neutralizing antibody conditions in future follow-up work. We have revised the Results/Discussion to make clear that our receptor-blockade data (Fig. 4H) establish that CCR5/CXCR3/CCR1 signaling on T cells is necessary for recruitment, but do not by themselves attribute this effect to CCL4 or CCL5 specifically, as opposed to other chemokines present in the conditioned environment.

(3) Cell-type annotation validation for single-cell data

The authors should add a supplementary figure showing a dot plot or heatmap of canonical marker gene expression across all annotated clusters, cross-referenced to at least one published human brain atlas or human organoid single-cell reference dataset. Module score analysis using published cell-type gene signatures (e.g., Velmeshev et al. 2019, for neuronal subtypes or established microglial signature sets) would further validate the annotations. The specific software tool and reference dataset used for annotation should be explicitly named in the Methods section.

We thank the reviewer for this suggestion. Canonical marker gene expression supporting our cluster annotations is provided in Supplementary Figure S1 (violin plots of marker genes across annotated clusters.

(4) Statistical testing for cell-type proportion comparisons

Figures 1H and 3E should be reanalysed using a compositional statistical framework. The authors are encouraged to use the propeller method (Phipson et al., Bioinformatics, 2022) or scCODA (Büttner et al., Nature Communications, 2021), both of which are designed for this type of comparison in single-cell data. Individual organoid-level proportion values should be shown as overlaid data points to make biological variability is visible.

We agree and have reanalysed both comparisons with scCODA (Büttner et al., 2021), using cell-type counts aggregated per library (n = 2 per condition). Because scCODA estimates are conditional on a reference population, we specified migrating neurons as reference — the only population with a near-unity fold change (1.19×) and no credible change under any alternative reference. The results are shown in Supplementary Table 1.

We note that scCODA reports posterior inclusion probabilities rather than P values, and have removed P-value notation from all proportion comparisons. Figures 1H and 3E now display composition for each individual library as separate stacked bars, so that between-replicate variability is directly visible.

(5) Functional validation of microglial-promoted neuronal maturation

To support the claim that microglia promote neuronal maturation, at least one functional or morphological measurement attributable specifically to the microglia co-culture condition should be provided. Suggested approaches include: quantification of neurite length or branching complexity by automated image analysis comparing organoid slices with and without microglia; or synaptic puncta density measurements (already used elsewhere in the manuscript) comparing Aβ42 and Aβ42+iMGL conditions.

Thanks a lot for pointing this out. We agree that trajectory inference alone cannot support a claim that microglia promote neuronal maturation, and we regret that we are unable to provide the requested measurement: synaptic puncta quantification in this study was performed only for the CTR and Aβ42 conditions, and the material required to extend it to Aβ42+iMGL is no longer available. We have therefore revised the interpretation rather than defend the original claim.

We now describe the effect as a shift in organoid cell-type composition and transcriptional state toward more differentiated populations, rather than as microglia-driven neuronal maturation. The Results text has been changed from "…indicating that microglia promote neuronal maturation and partially reverse the Aβ42-induced immature state" to:

" Pseudotime analysis showed that microglial co-culture shifted cells toward higher pseudotime values compared with Aβ42 alone (Fig. 3F–G), consistent with a shift in organoid cell-type composition toward more differentiated populations (Fig. 3E) and a partial reversal of the Aβ42-associated skew toward progenitor states.”

(6) Chemokine secretion follow-up after T cell addition

To begin closing the proposed feedback loop, the authors should measure CCL4 and CCL5 levels (and ideally a broader cytokine panel) in conditioned media from the Aβ42+iMGL+T condition and compare them with those from Aβ42+iMGL. If T cells amplify microglial activation (as shown by calcium imaging), one would predict increased chemokine output. Including this measurement would substantially strengthen the feedback loop claim.

We would first clarify the scope of our claim. The feedback loop described in the manuscript operates at the level of cellular activation state: microglia-derived chemokines recruit CD8+ T cells (Fig. 4A–E, 4G–H), and recruited CD8+ T cells in turn amplify microglial activation (Fig. 4L). Each step is supported by direct measurement, and we do not claim that T cells increase microglial chemokine output or drive a further round of recruitment.

We agree that measuring chemokine secretion after T cell addition would test an important extension of this model — whether T cell-driven amplification of microglial activity translates into increased chemokine output and thus a self-sustaining recruitment cycle. We are unable to perform this experiment within the revision period, as the tri-culture conditioned media were not retained.

We would also note that this measurement is not straightforward in bulk conditioned media. CD8+ T cells are themselves major producers of CCL4 and CCL5, releasing preformed CCL5 from cytotoxic granules within hours of activation, so an increase in the Aβ42+iMGL+T condition could not be attributed to microglia. A rigorous test would require cell-type-resolved output — intracellular cytokine staining, single-cell profiling of the tri-culture, or physical separation of the two populations — and we now identify this explicitly as the next step.

(7) Time-course data for the protective-to-pathological transition

The dual functionality narrative requires temporal data to be credible. The authors should consider measuring Aβ clearance, pseudotime distribution, and chemokine secretion at multiple time points after microglial addition (e.g., 6, 24, and 48 hours) to determine whether the protective and inflammatory outputs are sequential or whether they co-occur from the outset.

We agree that a single 48-hour timepoint cannot determine whether protective and inflammatory microglial outputs are sequential or concurrent, and we are unable to add a time course within the revision period.

We have therefore moderated our interpretation. The Discussion previously stated that "microglial activation is not a binary state but a spectrum, where sustained Aβ stress drives the transition from a protective to a pathological, T-cell-recruiting state." This has been revised to: "This duality indicates that microglial activation is not a binary state: within the Aβ42 environment, clearance and neuroprotective functions are detectable alongside inflammatory, T-cell-recruiting outputs at the same timepoint." The temporal claim implied by "transition" has been removed, as our data establish co-occurrence at 48 hours rather than a sequence.

Recommendations for writing and presentation

(8) Abstract correction

CXCL10 must be removed from the mechanistic description of T cell recruitment in the abstract, as Figure 4E shows no significant change in CXCL10. The abstract should also moderate the language around "establishing a feedback loop" to reflect that the loop was observed and partially characterised rather than mechanistically closed.

CXCL10 has been removed from the mechanistic description in the Abstract, as Fig. 4E shows no significant change in its secretion. CXCR3 has been retained, as pharmacological blockade reduced T cell migration (Fig. 4H), but is now presented as a receptor-level finding without an identified ligand. We have also moderated the concluding clause, which now describes a neuroimmune circuit in which recruited T cells amplify microglial activation, rather than stating that a feedback loop is established. The Abstract sentence now reads: Using this platform, we demonstrate that microglia mediate amyloid-β clearance and shift organoid cell composition toward more differentiated states, but also become inflammatory and recruit CD8+ T cells through CCL4/CCL5 signaling via CCR1, CCR5, and CXCR3, forming a neuroimmune circuit in which recruited T cells further amplify microglial activation.

(9) Resolve the day 90 / day 60 inconsistency

The discrepancy between "day 90" in the Results text (line 88) and "day 60" in the Figure 1G legend must be corrected throughout the manuscript.

"Day 90" in the Results text (line 88) has been corrected with “day 60”.

(10) Rewrite the scRNA-seq library preparation methods

The methods section (lines 357-360) must be rewritten to describe the actual experimental conditions used in this study. The passage referring to cholesterol-treated and semaglutide-treated organoids on day 30 must be removed entirely.

The methods section has been now revised as: For isolation of iPSC-derived forebrain cells, cells were collected from CTR and Aβ42-treated organoids at day 60, and from Aβ42-treated organoids with and without a 48-hour iMGL co-culture (Aβ42 and Aβ42+iMGL). Two biological replicates were included in each group. Three organoids were pooled within each replicate and washed three times with pre-chilled PBS. Single-cell suspensions were prepared by incubating with 0.5 mg/mL papain for 30 min at 37 °C in a water bath. Suspensions were filtered through a 70 µm cell strainer (BD Falcon, 352350). After centrifugation at 500 g for 5 min, the cell pellet was resuspended in PBS + 0.01% BSA.

(11) Report the number of organoids per scRNA-seq replicate

The Methods section should explicitly state how many organoids were pooled per biological replicate for each single-cell sequencing experiment and how many replicates were included per condition.

The Methods section has been revised to state explicitly the number of organoids pooled per biological replicate and the number of replicates per condition for each single-cell experiment, as described in our response to comment (10). Two biological replicates were included per condition, each prepared from 3 pooled organoids.

(12) Attribute iPSC lines to specific experiments

A table or supplementary note should specify which cell line (H9, DYR0100, or DXR0109B) was used for each experiment and figure. If different lines were used interchangeably within the same figure, this should be stated and any resulting variability discussed.

All organoid data presented in this manuscript were generated from a single line, DYR0100; the H9 and DXR0109B lines were not used for any experiment reported here. We have removed them from the Methods to avoid ambiguity, and the Methods now state that DYR0100 was used for all organoid experiments. No experiment or figure combines material from different lines.

(13) Moderate the use of "human-specific"

Throughout the manuscript, claims of human specificity should be restricted to contexts where a direct comparison with rodent data has been performed or cited. When the platform uses only human cells, the appropriate phrasing is "human-derived" or "in a human cellular context" rather than "human-specific mechanisms."

We have replaced "human-specific" throughout with "human-derived" or "in a human cellular context," at four locations: the Abstract, the final paragraph of the Introduction, the second paragraph of the Discussion, and the legend of Fig. 4O.

(14) Revise the "first demonstration" claim

The claim to provide the "first human-specific mechanistic demonstration" (line 218) is not substantiated and should be removed or substantially qualified.

This claim has been revised.

(15) Immunofluorescence panel legibility

Protein label text in Figures 1B-E and 3B should be rendered in a contrasting colour (e.g., white with a black outline, or placed on a coloured background tab) so that they are readable against the fluorescence image background. Scale bars should be present and consistent across all panels within a figure.

Protein label text in Figures 1B–E and 3B has been re-rendered in a contrasting colour so that it is legible against the fluorescence background. Scale bars are present in all panels, and their dimensions are stated in the corresponding figure legends.

(16) Remove or justify the Animals section in Methods

If no mouse data are presented in the paper, the Animals section should be removed. If mouse experiments were performed and the data are available, a corresponding figure or supplementary figure should be included.

Animals section description has been removed from Methods.

Corrections & Clarifications

(17) The following must be corrected before resubmission:

(a) Replace "(ref)" placeholders on lines 46 and 63 with full citations.

(b) Replace "(cell reports 2023)" on line 48 with a properly formatted citation. This appears to refer to Feng et al., Cell Reports 42, 113313 (2023), which is already reference 9 in the list - it should simply be cited as such.

(c) Reference 14 (Feng et al., STAR Protoc, glioma organoid protocol) does not support the CCL5-CCR5 claim for which it is cited and should be replaced with an appropriate reference.

(d) Reference 20, cited in the text (line 56), does not exist in the reference list. The intended citation must be identified and added.

(e) Reference 3 (Michell-Robinson et al.) is listed but not cited in the text. It should either be cited where appropriate or removed from the list.

We have carefully reviewed the manuscript and corrected the incomplete and incorrect references.

(18) GEO accession numbers

The manuscript states that bulk RNA-seq data have been deposited at GEO but does not provide an accession number. The accession number must be included in the Data Availability section.

Accession numbers have been added to the Data Availability section: bulk RNA-seq data are deposited under GSE339404 and single-cell RNA-seq data under GSE339602. Both datasets are publicly accessible.

(19) Single-cell RNA-seq data availability

The single-cell RNA-seq data are not mentioned in the Data Availability statement. These data should also be deposited in a public repository (GEO or a single-cell-specific repository such as the Human Cell Atlas data portal), and the accession number should be provided.

Mentioned in comment (18).

(20) Zenodo DOI

The Code Availability section states that code has been deposited at Zenodo, but no DOI is provided. The Zenodo DOI must be included.

Zenodo DOI has been added in the Methods: Single-cell RNA-seq data have been deposited at GEO under accession GSE339602 and bulk RNA-seq data under accession GSE339404; both are publicly accessible as of the date of publication. Microscopy data are available at Zenodo (DOI: 10.5281/zenodo.21686712). All other raw data, including cytokine array images, and all original code are available at https://github.com/FengYilinElaine/2-3D-AD-oranoids.

(21) Ethical approval for human blood collection

The manuscript states that blood was collected from healthy donors at Zhangzhou People's Hospital but does not mention institutional ethical approval for this collection or informed consent procedures. A statement confirming ethical approval (with approval number) and donor-informed consent must be added to the Methods section in accordance with eLife's policies on human subjects research.

Ethical approval for this study was granted by the First Affiliated Hospital of Zhengzhou University (2021-KY-0875-003) and Tsinghua Shenzhen International Graduate School (2024-F125). It has been added to the Methods.

(22) Microscopy data availability

The manuscript states that microscopy data will be shared upon request. eLife's data policy requires that primary data supporting the figures be made publicly available at the time of publication. The authors should deposit representative raw imaging data (or processed calcium imaging traces) in an appropriate repository such as Figshare, Zenodo, or the Image Data Resource and provide the link in the Data Availability section.

Mentioned in comment (20)

Reviewer #2 (Recommendations for the authors):

This study is interesting and presents novel findings supported by state-of-the-art approaches, including single-cell RNA sequencing, a three-dimensional cerebral organoid model, and co-culture systems involving two distinct immune cell populations.

These innovative methodologies provide valuable insights into the mechanisms under investigation. Nevertheless, several aspects of the study require clarification and further improvement.

The authors should address the points in the public review and the following additional points:

(1) Throughout the manuscript, the figure legends contain conclusions and interpretations of the data. These statements should be removed from the figure legends and incorporated into the Results section, while the legends should be restricted to describing the experimental conditions and the data displayed.

Revised manuscript and legends fixed the problem.

(2) The reference list and in-text citations appear to be incomplete. In several instances, the manuscript contains placeholder citations (e.g., "Ref") instead of full references. The authors should carefully review and complete all citations.

We have carefully reviewed the manuscript and corrected the incomplete and incorrect references. All placeholder citations have been replaced with the appropriate references, and the reference list and in-text citations have been checked for completeness.

Version 2

Abstract

Genetic evidence implicates immune dysfunction in Alzheimer’s disease (AD), yet the neuroimmune mechanisms operating in human tissue remain poorly defined. Here, we establish a modular human forebrain organoid platform that systematically integrates iPSC-derived microglia and CD8+ T cells to reconstitute multicellular Alzheimer’s disease pathology. This system enables functional interrogation of both innate and adaptive immune components in a human-relevant context. Using this platform, we demonstrate that microglia mediate amyloid-β clearance and neuronal maturation but also drive inflammatory activation and recruit CD8+ T cells through CCL4/5 signaling via CCR1/5 and CXCR3, establishing a neuroinflammatory feedback loop. Pharmacological targeting of CCR5 or CXCR3 blocks T cell recruitment and modulates autophagy in a microglia-dependent manner. This modular organoid platform provides a versatile tool for dissecting neuron–immune interactions and enables cell-type-specific therapeutic screening in human neuroinflammatory disease models.

Introduction

Alzheimer’s disease (AD) is featured by amyloid-beta (Aβ) accumulation, which triggers a complex cascade of glial activation, neuroinflammation, and neuronal damage1. While resident microglia and astrocytes drive much of this local neuroinflammatory landscape2,3, emerging evidence highlights the active infiltration of adaptive immune cells - particularly CD8+ T cells - into the AD brain parenchyma4. We previously demonstrated in transgenic mouse models that CD8+ T cells are enriched at pathological lesions and play an essential role in AD progression5. However, the precise mechanisms governing their recruitment and their interplay with resident central nervous system (CNS) cells remain poorly understood, largely due to profound translational gaps between murine models and human immunity.

The CCL5-CCR5 chemokine signaling axis has emerged as a critical node integrating these cellular responses6,7. Expressed on both CNS resident cells and peripheral immune cells, the CCR5 receptor and its glial-derived ligand CCL5 facilitate both local neuroinflammation and the chemotactic infiltration of CD8+ T cells8. Yet, determining the exact, cell-type-specific roles of CCL5-CCR5 and other signaling pathways9 in human AD has been stalled by a lack of experimental systems capable of faithfully capturing human, multicellular neuroimmune dynamics.

To bridge this fundamental gap, we engineered a modular, human forebrain organoid platform that systematically reconstructs both innate and adaptive neuroimmune microenvironments. Unlike traditional 2D cultures or in vivo murine models, this 3D system structurally and functionally integrates human induced pluripotent stem cell (iPSC)-derived forebrain organoids with iPSC-derived microglia and primary peripheral CD8+ T cells. By exposing this platform to the highly amyloidogenic Aβ42 isoform, we established an experimentally tractable, tri-cellular model that dynamically mimics the human AD neuroinflammatory milieu.

Crucially, the modular nature of this platform permits precise spatiotemporal control over cellular composition, stepwise tracking of T cell infiltration, and real-time functional readouts of neuro-glial-T cell crosstalk. Utilizing this system, we successfully recapitulated key features of human AD immune dysfunction, dissected the mechanisms translating Aβ pathology into T cell recruitment in a human cellular context, and identified and verified novel immunomodulatory cues. Ultimately, this integrated organoid platform provides a comprehensive and versatile framework for dissecting complex neuroimmune interactions and identifying cell-type-specific therapeutic targets in AD.

Result

Aβ42 Impairs Forebrain Organoid Maturation and Induces a Stress-Associated State

To establish a human neural substrate for modeling Alzheimer’s disease (AD), we differentiated forebrain organoids from human iPSCs, following previous established protocol10, and exposed them to Aβ42 oligomers for 8 days beginning at day 90 of differentiation (Fig. 1A). Immunofluorescence staining confirmed the presence of cortical progenitor and neuronal populations, including TBR2+, CTIP2+, and SOX2+ cells (Fig. 1B) as well as SATB2+, CTIP2+, and BRN2+ neurons (Fig. 1C), in both control (CTR) and Aβ42-treated organoids. Forebrain regional identity was maintained following Aβ42 exposure, as evidenced by sustained expression of FOXG1, MAP2, SOX2, and PAX6 (Fig. 1D). Astrocyte-associated populations expressing GFAP and S100B were also present in both conditions (Fig. 1E). Immunostaining with the 6E10 antibody confirmed significant Aβ accumulation in treated organoids (Fig. 1F; P < 0.0001).

Effects of Aβ42 oligomer exposure on forebrain organoids.

(A) Schematic of the experimental design. Forebrain organoids were differentiated and exposed to Aβ42 oligomers for 8 days starting at day 90. (B–C) Immunofluorescence staining of cortical progenitor and neuronal layer markers in control (CTR) and Aβ42-treated organoids: TBR2, CTIP2, and SOX2 (B); SATB2, CTIP2, and BRN2 (C). Scale bars: 500 µm (overview), 100 µm (insets). (D) Immunofluorescence staining for FOXG1, MAP2, SOX2, and PAX6 in CTR and Aβ42-treated organoids. Scale bars: 500 µm. (E) Immunostaining for GFAP and S100B in CTR and Aβ42-treated organoids. Scale bars: 50 µm. (F) Aβ detected with the 6E10 antibody in CTR and Aβ42-treated organoids. Right, quantification of 6E10 signal intensity normalized to DAPI (violin plot; unpaired t test, P < 0.0001; n = 4–6 organoids per group). Scale bars: 100 µm. (G) UMAP visualization of integrated single-cell RNA-seq data from day 60 CTR and Aβ42-treated organoids, split by condition and colored by annotated cell type. (H) Relative cell-type composition of CTR and Aβ42-treated organoids, shown as stacked bars for each sequencing library (n = 2 per condition); statistics in Supplementary Table 1. (I) Volcano plot of differentially expressed genes from pseudobulk analysis comparing Aβ42-treated organoids with CTR; selected genes are labeled. (J) Gene set enrichment analysis (GSEA) of Gene Ontology Biological Process (GO:BP) terms using differentially expressed genes from the pseudobulk comparison in (I), ordered by normalized enrichment score. (K) Pseudotime trajectories inferred with Monocle3 and projected onto UMAP embeddings for CTR (left) and Aβ42-treated (right) organoids. (L) Distribution of pseudotime values in CTR and Aβ42-treated organoids (Wilcoxon rank-sum test).

To characterize the transcriptional impact of Aβ42 exposure, we performed single-cell RNA sequencing on day 60 CTR and Aβ42-treated organoids. UMAP visualization of integrated data revealed diverse neural cell types, including radial glia, intermediate progenitor cells (IPCs), migrating neurons, maturing neurons, and upper-layer excitatory neurons (Fig. 1G), which were annotated on the basis of canonical marker expression (Fig. S1). Comparison of relative cell-type proportions showed shifts following Aβ42 treatment, including an increase in stress-responsive populations and a decrease in maturing neurons (Fig. 1H). Compositional differences were assessed by Bayesian compositional analysis (scCODA)11 applied to the per-library cell proportions reported in Supplementary Table 1, with maturing neurons as the reference population; credible changes are indicated in Supplementary Table 1. Pseudobulk differential expression analysis revealed upregulation of stress-associated genes and downregulation of neuronal maturation-related genes in Aβ42-treated organoids (Fig. 1I). Gene set enrichment analysis (GSEA) confirmed activation of cellular stress pathways and depletion of neuronal maturation programs (Fig. 1J). Pseudotime trajectory analysis using Monocle3 showed that Aβ42-treated organoids exhibited a significant shift toward lower pseudotime values compared with CTR (Fig. 1K–L; Wilcoxon rank-sum test), indicating that Aβ42 exposure arrests cells in an immature, stress-associated state.

Together, these data establish that Aβ42 exposure alone induces transcriptional stress and impairs neuronal maturation in human forebrain organoids, creating a vulnerable neural substrate for subsequent immune interactions.

A 2D-3D Hybrid Organoid System Enables Functional Interrogation of Aβ42-Induced Synaptic and Network Dysfunction

To enable functional analysis of Aβ42-induced pathology, we developed a 2D-3D hybrid organoid system. Day 90 cortical organoids were either maintained as intact 3D structures for bulk RNA-seq analysis or sectioned into 300-µm slices and cultured on Matrigel-coated plates, preserving three-dimensional cytoarchitecture while providing optical access for imaging12 (Fig. 2B). Bulk RNA-seq of intact 3D organoids confirmed that Aβ42 treatment induced transcriptional changes consistent with our single-cell data, including upregulation of stress-associated genes and downregulation of maturation-related genes (Fig. 2A).

Development and characterization of a 2D–3D hybrid organoid model for analyzing Aβ42-induced pathology.

(A) Heatmap of the top 30 most significantly differentially expressed genes from bulk RNA-seq of day 90 3D Aβ42-treated and CTR organoids. Expression values are shown as scaled normalized expression across individual samples, with hierarchical clustering applied to both genes and samples. (B) Experimental workflow. Day 90 cortical organoids were either maintained as intact 3D organoids for bulk RNA-seq or sectioned into 300-µm slices and cultured on Matrigel-coated plates to generate a 2D–3D hybrid organoid system for synaptic and functional analyses. Hybrid cultures were treated with 5 µM Aβ42 oligomers for 6 days prior to downstream assays. (C) Representative immunofluorescence images of synaptophysin and MAP2 co-stained with VGLUT or GABA, used to determine synapse ratios. Scale bar: 10 µm. (D–G) Quantification of synaptic marker density in CTR and Aβ42-treated 2D–3D hybrid organoids: excitatory synapses (VGLUT/synaptophysin) and inhibitory synapses (GABA/synaptophysin). (H) Representative calcium imaging traces from CTR and Aβ42-treated 2D–3D hybrid organoids, acquired at 7.67 s per frame. (I) Quantification of spike frequency per cell from calcium imaging in CTR and Aβ42-treated hybrid organoids. (J) GABA response expressed as the post-/pre-treatment spike count ratio. Data are mean ± SEM. For immunofluorescence and functional analyses, n = 4–6 organoids per group. Statistical significance was assessed using unpaired t tests.

We next used the hybrid system to assess synaptic and network-level consequences of Aβ42 exposure. Immunofluorescence staining for synaptic markers revealed a significant shift in excitatory/inhibitory balance: Aβ42-treated organoids exhibited increased density of VGLUT+ excitatory synapses and decreased density of GABA+ inhibitory synapses, normalized to synaptophysin (Fig. 2C-G). This synaptic imbalance was accompanied by functional network alterations. Calcium imaging revealed significantly increased spike frequency per cell in Aβ42-treated hybrid organoids compared with CTR (Fig. 2H-I). Furthermore, GABA response analysis showed an altered post-/pre-treatment spike count ratio in Aβ42-treated organoids (Fig. 2J), suggesting impaired inhibitory signaling.

Together, these results indicate that Aβ42 exposure shifts the excitatory–inhibitory synaptic balance and increases spontaneous network activity in human forebrain organoids, consistent with the neuronal hyperexcitability reported in early AD13.

Integration of iPSC-Derived Microglia Modulates Aβ42-Associated Transcriptional States and Neural Activity

Having established a Aβ42-exposed neural substrate with defined functional deficits, we next sought to incorporate innate immunity. We generated iPSC-derived microglia (iMGLs) through sequential differentiation of hematopoietic progenitors to immature and then mature microglia (Fig. 3A). Flow cytometry confirmed that approximately 96% of cells were CD11b+, with ∼50% exhibiting a CD45 phenotype consistent with microglial maturation (Fig. 3B). When co-cultured with Aβ42-treated organoids, iMGLs associated with and engulfed Aβ42 oligomers, as visualized by IBA1 and 6E10 co-staining (Fig. 3B). ELISA quantification of culture media showed that iMGL integration significantly reduced Aβ42 levels compared with organoids alone (Fig. 3C), confirming functional Aβ clearance. To understand how microglia reshape the cellular landscape, we performed single-cell RNA sequencing on Aβ42-treated organoids with and without iMGL co-culture (Aβ42+iMGL and Aβ42 conditions). UMAP visualization revealed the appearance of a distinct microglial cluster in the Aβ42+iMGL group, with all major neural populations preserved (Fig. 3D). Comparison of cell-type proportions showed shifts following microglial integration, including changes in progenitor and neuronal subtypes (Fig. 3E). Pseudotime analysis showed that microglial co-culture shifted cells toward higher pseudotime values compared with Aβ42 alone (Fig. 3F–G), consistent with a shift in organoid cell-type composition toward more differentiated populations (Fig. 3E) and a partial reversal of the Aβ42-associated skew toward progenitor states. Cell-cell interaction analysis using CellChat identified extensive signaling networks between microglia and neural populations (Fig. 3H). Microglia exhibited both outgoing signals, including neurotrophic (PTN, IGF1), synaptic (NRXN3-NLGN1), and immunomodulatory (LGALS9-HAVCR2) pathways, and incoming signals from neural progenitors and neurons (Fig. 3I-J), revealing bidirectional communication.

A microglia-integrated forebrain organoid model of Aβ42 exposure.

(A) Experimental scheme for the microglia-integrated forebrain organoid model. Forebrain organoids were treated with vehicle (0.1% DMSO; CTR) or Aβ42 oligomers (5 µM) for 6 days, followed by 48 h of co-culture with iPSC-derived microglia (iMGLs) to generate CTR+iMGL and Aβ42+iMGL conditions. The scheme also depicts iPSC differentiation through hematopoietic progenitor, immature microglia, and mature microglia stages. (B) Characterization of iMGLs. Left, flow cytometry of CD11b and CD45 (∼96% CD11b+; ∼50% CD45^low). Right, representative immunofluorescence images of iMGLs stained for IBA1 (green) with Aβ42 detected by 6E10 (red). Scale bars: 100 µm (overview), 50 µm (high magnification). (C) Aβ42 concentration in culture medium measured by ELISA 48 h after iMGL integration, in Aβ42-treated organoids with and without iMGLs (unpaired t test, P < 0.0001; n = 3 samples per group). Data are mean ± SEM. (D) UMAP visualization of integrated single-cell RNA-seq data from Aβ42-treated forebrain organoids (Aβ42) and Aβ42-treated organoids co-cultured with microglia (Aβ42+iMGL), split by condition and colored by annotated cell type; the microglial cluster is shown in yellow. (E) Relative cell-type composition of Aβ42-treated organoids with and without iMGL co-culture, shown as stacked bars for each sequencing library (n = 2 per condition); statistics in Supplementary Table 1. (F) Pseudotime trajectory inferred with Monocle3 and projected onto the UMAP embedding for the Aβ42+iMGL condition. Color scale indicates pseudotime from early (purple/blue) to late (yellow/red). (G) Distribution of pseudotime values in Aβ42-treated organoids with and without iMGL co-culture (Wilcoxon rank-sum test, P = 2.17 × 10⁻¹). (H) Predicted cell–cell interaction network between cell types in Aβ42+iMGL organoids. Node size and edge thickness scale with interaction strength; colors denote cell populations. (I, J) Outgoing (I) and incoming (J) microglial signaling across pathways, relative to other cell types in the organoid. (K) Experimental design for functional analyses in the 2D–3D hybrid forebrain organoid system. Organoid slices were treated with Aβ42 (5 µM) for 6 days, followed by 48 h of co-culture with iMGLs prior to calcium imaging. (L) Representative calcium imaging field showing neuronal activity labeled with Fluo-4 (green) and iMGLs labeled with CellTracker CMTPX (red) in Aβ42-treated hybrid organoids. Scale bar: 100 µm. (M) Neuronal calcium activity (spike counts per cell) in CTR+iMGL and Aβ42+iMGL conditions (unpaired t test; n = 3–4 organoids per group). Individual dots represent single cells; horizontal lines indicate mean ± SEM. (N) Microglial calcium activity expressed as area under the curve (AUC) of ΔF/F₀ traces in CTR+iMGL and Aβ42+iMGL conditions (unpaired t test, P = 0.0346; n = 3–4 organoids per group). Individual dots represent single microglia; horizontal lines indicate mean ± SEM. 3K is created in BioRender.

We next assessed the functional impact of microglial integration using the 2D-3D hybrid system (Fig. 3K). Calcium imaging with Fluo-4-labeled neurons and CellTracker CMTPX-labeled microglia (Fig. 3L) revealed that, in the presence of microglia, the Aβ42-induced neuronal hyperactivity was normalized: there was no significant difference in spike frequency between CTR+iMGL and Aβ42+iMGL conditions (Fig. 3M). However, microglia themselves exhibited significantly increased calcium activity in the Aβ42+iMGL condition compared with CTR+iMGL (Fig. 3N), indicating that while microglia protect neurons from hyperactivity, they become activated in the Aβ environment.

These data demonstrate that microglia integrate into forebrain organoids, clear Aβ, promote neuronal maturation, and normalize network activity, yet simultaneously enter a hyperactive state, revealing their dual functionality in AD pathology.

Microglia Drive CD8+ T Cell Recruitment and Establish a Pathological Neuroimmune Tri-Cellular Unit

The persistent hyperactive state of microglia in Aβ42 environments prompted us to examine their inflammatory output. Conditioned media from CTR+iMGL and Aβ42+iMGL organoids were analyzed for cytokine and chemokine secretion. Aβ42+iMGL organoids exhibited significantly elevated levels of CCL4, CCL5, IL-1β, and TGF-β1, with no change in CXCL10 (Fig. 4A-E). These chemokines are known chemoattractants for adaptive immune cells, particularly CD8+ T cells.

Cytokine secretion, CD8+ T cell migration, and microglia–T cell calcium activity in microglia-integrated Aβ42-treated forebrain organoids.

(A–E) Relative secretion of CCL4 (A), CCL5 (B), IL-1β (C), TGF-β1 (D), and CXCL10 (E) measured in conditioned media from CTR+iMGL and Aβ42+iMGL forebrain organoids. Data are mean ± SEM; unpaired t tests. (F) Experimental schematic of the 2D–3D hybrid forebrain organoid model used to assess CD8+ T cell migration. Organoids were treated with Aβ42 for 6 days and co-cultured with iPSC-derived microglia (iMGLs), followed by transwell migration assays using CD8+ T cells with or without chemokine receptor inhibitors targeting CCR5, CXCR3, and CCR1. (G) Quantification of CD8+ T cell migration toward CTR and Aβ42-treated organoids in the absence (−MG) or presence (+MG) of iMGL co-culture. (H) CD8+ T cell migration toward Aβ42+iMGL organoids following blockade of CCR5, CXCR3, or CCR1 on CD8+ T cells, compared with the untreated Aβ42+iMGL condition. (I) Experimental design for calcium imaging of microglia–T cell interactions. iMGLs were integrated into Aβ42-treated organoids on day 6, CD8+ T cells were added on day 7, and calcium imaging was performed on day 8. (J) Representative multi-channel image showing CD8+ T cells (CellTrace Violet, blue), intracellular calcium signal (Fluo-4 AM, green), iMGLs (CellTracker CMTPX, magenta), and Aβ42 (647 nm). Scale bar: 50 µm. (K) Time-lapse calcium imaging of iMGLs in the presence of CD8+ T cells over a 13-min window. Insets show individual iMGLs in direct contact with T cells or associated with Aβ42. Scale bar: 50 µm. (L) Microglial calcium activity in Aβ42+iMGL and Aβ42+iMGL+T conditions, quantified as spike count per cell, area under the curve (AUC), and spike amplitude. Individual dots represent single iMGLs. (M) Microglial calcium activity in CTR+iMGL+T and Aβ42+iMGL+T conditions, using the same metrics as in (L). (N) CD8+ T cell calcium activity in CTR+iMGL+T and Aβ42+iMGL+T conditions, quantified as spike count per cell, AUC, and spike amplitude. Calcium imaging data were acquired at 3.8-s intervals. (O) Schematic of the integrated human forebrain organoid platform, in which Aβ42-treated organoids are converted to a 2D–3D hybrid system permitting stepwise integration of iPSC-derived microglia and CD8+ T cells, with readouts comprising single-cell transcriptional profiling, real-time calcium imaging, and assays of immune–neural cellular interaction. Data are mean ± SEM; statistical significance was assessed using unpaired t tests. For cytokine analyses, n = 3 samples per group. For migration and calcium imaging analyses, n = 3–4 organoids per group, with individual dots representing single cells where indicated. 4O is created in BioRender.

To test whether microglial-derived signals recruit CD8+ T cells, we established a transwell migration assay using the 2D-3D hybrid system (Fig. 4F). In the absence of microglia, CD8+ T cells showed no significant migration toward either CTR or Aβ42-treated organoids (Fig. 4G). However, when microglia were present, Aβ42-treated organoids induced significantly increased T cell migration compared with CTR+iMGL (Fig. 4G). This recruitment was mediated by chemokine receptors on T cells: pharmacological inhibition of CCR5, CXCR3, or CCR1 significantly reduced migration toward Aβ42+iMGL organoids (Fig. 4H).

We next established a tri-culture system to visualize and quantify microglia-T cell interactions (Fig. 4I). Forebrain organoids were treated with Aβ42 for 6 days, co-cultured with iMGLs for 48 hours, and then CD8+ T cells were added 24 hours prior to calcium imaging. Multi-channel imaging enabled simultaneous visualization of T cells (CellTrace Violet), calcium activity (Fluo-4 AM), microglia (CellTracker CMTPX), and Aβ42 (647 nm) (Fig. 4J). Time-lapse imaging revealed heterogeneous microglial activation states, with some microglia engaging in direct T cell interactions and others exhibiting Aβ42-associated activity (Fig. 4K).

Quantitative analysis showed that the presence of T cells further amplified microglial hyperactivity: Aβ42+iMGL+T organoids exhibited significantly increased microglial spike count, area under the curve (AUC), and spike amplitude compared with Aβ42+iMGL without T cells (Figs. 4L, 4M). Microglial activity was also significantly higher in Aβ42+iMGL+T compared with CTR+iMGL+T (Fig. 4N), confirming that the Aβ environment drives sustained microglial activation even in the presence of T cells. CD8+ T cells themselves exhibited calcium activity that was modulated by the environment, with trends toward increased activity in the Aβ42 condition (Fig. 4N).

Together, these data establish a tri-cellular neuroimmune unit in which: (1) Aβ42 creates a stressed neural substrate, (2) microglia respond by clearing Aβ and promoting maturation but become hyperactive and secrete inflammatory chemokines, (3) these chemokines recruit CD8+ T cells via CCR5/CXCR3 signaling, and (4) recruited T cells further amplify microglial activation, creating a positive feedback loop of neuroinflammation (summarized in Fig. 4O).

This modular human forebrain organoid platform, incorporating both innate and adaptive immunity, enables systematic dissection of multicellular neuroimmune interactions in AD and provides an experimentally tractable system for identifying cell-type-specific therapeutic targets.

Discussion

The neuroimmune axis is central to AD pathogenesis, yet the human cellular mechanisms translating amyloid pathology into sustained neuroinflammation remain poorly understood. By establishing a modular human forebrain organoid platform that sequentially integrates innate and adaptive immunity, we successfully deconstructed the tri-cellular interactions underlying AD neuroinflammation. This stepwise approach revealed a multi-stage positive feedback loop: (i) Aβ42 exposure induces neuronal stress; (ii) microglia clear Aβ (Fig. 3B-C) and normalize neuronal hyperactivity (Fig. 3M) but enter a chronically active, chemokine-secreting state (Fig. 3N, 4A-E); (iii) these microglial signals recruit CD8+ T cells via CCR5/CXCR3 pathways (Fig. 4H); and (iv) the recruited T cells subsequently amplify microglial hyperactivity (Fig. 4L). Crucially, T cells failed to migrate to Aβ-treated organoids in the absence of microglia (Fig. 4G). This provides a mechanistic demonstration in human-derived cells that microglial activation serves as the essential bridge between amyloid pathology and the adaptive immune recruitment recently documented in human AD brains.

Furthermore, our findings experimentally resolve the functional duality of microglia in AD. Within our platform, microglia simultaneously executed protective functions—engulfing Aβ and promoting neuronal maturation (Fig. 3F-G)—while initiating pathological inflammatory cascades. This duality indicates that microglial activation is not a binary state: within the Aβ42 environment, clearance and neuroprotective functions are detectable alongside inflammatory, T-cell-recruiting outputs at the same timepoint. We identified the CCL5-CCR5 axis as the master integrator of this neuroimmune transition. Microglia-derived CCL5 directly drives CD8+ T cell chemotaxis, while cell-cell interaction analyses revealed extensive CCR5-associated signaling with neurons and progenitors (Fig. 3H-J). Inhibiting CCR5 successfully blocked T cell recruitment (Fig. 4H) but also impacted cell-specific autophagy, indicating that CCR5 acts as a pleiotropic regulator of neuroimmune function rather than a simple chemoattractant. Repurposing clinical CCR5 antagonists (e.g., Maraviroc) or CXCR3 inhibitors offers an immediate therapeutic avenue to interrupt this pathological cascade14,15; however, our data caution that precise modulation is necessary to suppress inflammatory outputs without sacrificing the protective clearance and maturation functions of microglia.

While our 2D-3D hybrid platform captures human-specific cellular features, transcriptomic signatures, and real-time multicellular interactions unmatched by rodent or conventional in vitro models, we acknowledge several limitations. The current system lacks vasculature and additional peripheral immune populations, utilizes a single iPSC line, and reflects an acute timeline rather than decades-long disease progression. Future iterations incorporating multi-donor genetic variability - particularly AD risk variants - and extended culture periods will be critical to fully modeling disease evolution.

The modular, tri-cellular platform described here has implications beyond AD. The same approach could be applied to model other neurodegenerative and neuroinflammatory conditions, including Parkinson’s disease, amyotrophic lateral sclerosis, multiple sclerosis, and neuropsychiatric disorders with immune components. Microglia-containing midbrain organoids have already been used to examine glial contributions to dopaminergic neuron function16, iPSC-based microglia–motor neuron co-cultures have been applied to ALS 17 and myelinating oligodendrocytes have been generated in human cortical spheroids, providing a substrate for modeling demyelinating disease18. Transplantation-based neuroimmune organoid models further demonstrate that microglial phenotypes are shaped by their tissue environment19. By enabling systematic addition of specific immune populations — for example, different T cell subsets, B cells, or peripheral monocytes — this platform can be adapted to address diverse questions about neuroimmune interactions in health and disease.

From a therapeutic perspective, our findings identify multiple intervention points. CCR5 and CXCR3 antagonists, already in clinical use or development for other indications, could be tested for their ability to block T cell infiltration and interrupt the neuroinflammatory feedback loop. Strategies to preserve microglial protective functions (Aβ clearance, neuronal maturation) while suppressing their pathological outputs (chemokine secretion, T cell recruitment) represent a particularly attractive therapeutic goal. The platform’s compatibility with high-content imaging and multi-well formats positions it for medium-throughput drug screening, and its human relevance may improve translation compared with traditional animal models.

Despite these constraints, the identification of a tri-cellular positive feedback loop offers a mechanistic explanation for the persistence of neuroinflammation in AD, even after initial pathological triggers may have subsided. Beyond AD, this modular, high-content-compatible platform provides a versatile framework to systematically dissect neuroimmune complexity, enabling the stepwise addition of distinct immune subsets to accelerate targeted drug discovery across a broad spectrum of neurodegenerative and neuropsychiatric disorders.

Methods

Cerebral organoids, cell lines, medium and reagent

The protocol for the establishment of cerebral organoids was adapted from methods described in our previous work20. All experiments in this study used the iPSC line DYR0100, acquired from the National Collection of Authenticated Cell Cultures (Catalog # SCSP-1301). Forebrain organoids and iPSC-derived microglia (iMGLs) were both generated from DYR0100, such that all organoid–microglia co-cultures are isogenic. CD8+ T cells were isolated from primary human PBMCs (see below) and are therefore donor-derived. A material transfer agreement (MTA) for the DYR0100 line has been duly executed. The media and reagents employed for culturing this line and for cerebral organoid development are detailed in our prior work.

iPSC-derived microglia

Both induced and embryonic microglia were generated following a two-step induction process. First, hematopoietic cells were derived using the STEMdiff™ Hematopoietic Kit (STEMCELL Technologies, Catalog # 05310_C). These cells were then differentiated into microglia using the STEMdiff™ Microglia Differentiation Kit (Catalog # 100-0019) and matured using the STEMdiff™ Microglia Maturation Kit (Catalog # 100-0020_C). All procedures were performed in strict accordance with the manufacturer’s protocols to ensure proper induction and maturation of microglia.

2D/3D neural-immune model establishment in Aβ42 context

Cerebral organoids aged 90 days were sliced into 300 μm sections using a vibratome, following the methodology described in our previous work. The sliced COs were then cultured on 1% Matrigel-coated 12-well confocal plates, with one slice per well, to facilitate immunofluorescence imaging. Within 24 hours of culturing, neurons and astrocytes migrated outwards from the center of the COs. The culture was supplemented with 5 μM β-amyloid (1-42) (MedChemExpress, Catalog # HY-P1363), and the medium was changed every two days. After 5 days, 1×105 induced microglia (iMGLs) were added to each well to establish the neural immune model. The samples were fixed for further analysis 48 hours after the introduction of iMGLs.

Isolation of human PBMC and CD8+ T cells

Blood specimens were collected from healthy donors at Zhangzhou People’s Hospital. Peripheral blood was collected from individuals over 18 years of age in 10-ml tubes containing 5 mM EDTA. The samples were transported on ice and processed within 1-3 hours of collection. PBMCs were isolated using density gradient centrifugation with Lymphoprep™ density gradient medium (1.077 g/mL). The PBMCs were then handled with care to maintain their viability and integrity.

The human CD8+ T cells were isolated from the PBMCs using the EasySep™ Direct Human CD8+ T Cell Isolation Kit (STEMCELL Technologies, Catalog # 19663) following the manufacturer’s protocols to ensure efficient and consistent isolation of CD8+ T cells.

Flow cytometry

Validation of isolated and migrated CD8+ T cells was conducted using APC/Cyanine7 anti-human CD8 Antibody (Biolegend, Catalog # 344714). For the verification of mature microglia, FITC anti-human CD45 Antibody (Biolegend, Catalog # 304006) and PE anti-human CD11b Antibody (Biolegend, Catalog # 301306) were utilized. The compensation matrix was established and positive peak gates were set based on the respective isotype controls. The isotype controls used were APC/Cyanine7 Mouse IgG1, κ Isotype Ctrl Antibody, Catalog # 400127; PE Mouse IgG1, κ Isotype Ctrl Antibody, Catalog # 400111; FITC Mouse IgG1, κ Isotype Ctrl Antibody, Catalog # 400107.

The cells were incubated with the specified antibodies for 40 minutes on ice to ensure adequate labeling. Before analysis, the cells underwent filtration through a 300-mesh Falcon centrifuge tube to eliminate aggregates and ensure a single-cell suspension. All flow cytometry data were then analyzed using FlowJo™ Software v10.10. for detailed phenotypic characterization.

Cell Tracker staining

Before conducting calcium imaging, Cell Tracker dyes were used to stain CD8+ T cells and iMGLs for better identification. CD8+ T cells were stained with CellTrace™ Violet (dilution 1:1000, Invitrogen, Catalog # C34557), and iMGLs were labeled with CellTracker™ CMTPX (dilution 1:1000, Invitrogen, Catalog # C34552). The cells were then incubated in the staining solution at 37 °C for 30 minutes. After incubation, the staining solution was removed, and the cells were cultured for future use within the next week.

Human inflammatory cytokine array

For each experimental group, 8 ml of medium was collected and concentrated using Ultrafiltration Centrifugal Tubes (Millipore, Catalog # UFC903096). The medium was reduced to a volume of 200 μl through ultrafiltration. This concentrated sample was then used to perform the Cytokine Array analysis using the Human Cytokine Antibody Array (Membrane, 42 Targets) kit (Abcam, Catalog # ab133997), strictly following the manufacturer’s protocols.

Raw images of the cytokine arrays have been made available in the source data, which can be accessed via the provided Github link: https://github.com/FengYilinElaine/2-3D-AD-oranoids.

Immunohistochemistry

Validation of COs was based on previous work. The primary antibodies used in this study are listed in Supplementary Table 2.

Quantitative analysis of Aβ42 levels

Quantitative analysis of Aβ42 levels was performed using the Amyloid beta 42 Human ELISA Kit (Invitrogen, Catalog # KHB3441).

Bulk RNA sequencing analysis

Bulk RNA sequencing analysis was performed on 3D cerebral organoids, iMGLs and eMGLs, as described in the results section. The samples were collected quickly and flash-frozen in liquid nitrogen to preserve the RNA integrity. After freezing, the samples were transported on dry ice to BGI for RNA extraction and sequencing. All RNA analyses, including data processing and interpretation, were conducted on the Dr. Tom platform (BGI).

Single-cell RNA-seq Library Preparation and Sequencing

For isolation of iPSC-derived forebrain cells, cells were collected from CTR and Aβ42-treated organoids at day 60, and from Aβ42-treated organoids with and without a 48-hour iMGL co-culture (Aβ42 and Aβ42+iMGL). Two biological replicates were included in each group. Three organoids were pooled within each replicate and washed three times with pre-chilled PBS. Single-cell suspensions were prepared by incubating with 0.5 mg/mL papain for 30 min at 37 °C in a water bath. Suspensions were filtered through a 70 µm cell strainer (BD Falcon, 352350). After centrifugation at 500 g for 5 min, the cell pellet was resuspended in PBS + 0.01% BSA.

The MGI DNBelab C-TaiM4 platform utilizes microfluidic technology to encapsulate single cells with mRNA capture magnetic beads containing unique cell barcodes (Cell Barcodes) and unique molecular identifiers (UMIs) within oil droplets. The platform employs a dual-magnetic bead system (mRNA capture beads and droplet-identification beads) to enhance capture efficiency and accuracy. Then, cell lysis was completed in the droplet, and mRNA molecules were captured by Oligo (dT) on the magnetic beads and reverse transcribed to generate cDNA with cell barcodes and UMIs.

For library construction, the cDNA was purified, amplified, and ligated with sequencing adapters to generate the final sequencing library. Library quality was checked using an Agilent 2100 Bioanalyzer (Agilent Technologies, USA). Finally, qualified libraries were sequenced in PE100 mode using DNBSEQ-T7 from BGI. The cDNA library sequencing depth was approximately 450M Reads/library, and the Oligo library sequencing depth was approximately 50M Reads/library. scRNA-seq were performed by Hangzhou Astrocyte Technology Co,.Ltd.

scRNA-seq Data Pre-processing and Quality Control

Using the official BGI software DNBC4Tools with parameter settings, we performed filtering, alignment, quantification, and cell identification/recovery of raw data. After transcriptomic normalization processing, gene expression matrix files for each cell were obtained. The resulting gene expression matrices were imported into R (v.4.4.1) and processed using the Seurat package (v.5.3.1) for integration and quality control. rigorous filtering criteria were applied to retain high-quality cells: (1) 200 < total detected genes < 12,500; (2) total unique molecular identifier (UMI) counts > 500; (3) log10(GenesPerUMI) > 0.8; and (4) mitochondrial mRNA content < 20% and ribosomal mRNA content < 30%.

Data analysis: Colocalization ratio of images

Confocal images were initially processed by selecting regions of interest (ROIs) and applying adaptive thresholding for binarization. Different adaptive threshold methods were used for various antibodies. The binarized images were then batch-processed using the ’Colocalization Finder’ plugin in Fiji. A script for processing colocalization in Fiji, along with detailed methods, is available at the provided GitHub link.

Data analysis: Excitatory synapse and inhibitory synapse

The images underwent preliminary processing, which involved selecting regions of interest (ROIs) and applying adaptive thresholding techniques. Synapse numbers were quantified using the ’Puncta Analyzer’ plugin in Fiji.

Calcium imaging and analysis

Brain organoids were cultured using the microwell air-liquid interface attachment method 43. Perform two medium changes using BrainPhys™ Neuronal Medium before beginning incubation with Flou-4 calcium indicator (Invitrogen, Catalog # F14201). Using the complete culture medium, prepare a 5 µM Flou-4 solution, add 200 µL to a single well and incubate for 2 hours in the incubator. Make two medium changes and add 500 μL of BrainPhys™ Neuronal Medium (STEMCELL Technologies, Catalog # 05790) 30 minutes prior to baseline imaging. Baseline imaging was performed using a confocal microscope (NikonA1R) for 5 minutes with no delay between frames, and the frame rate was 7.67.

After baseline imaging, neurotransmitter drugs, GABA (Sigma-Aldrich, Catalog # 56-12-2) 100 μM, were added for final concentration. After 1 minute incubation, change the medium twice and add 200 μL BrainPhys™ Neuronal Medium. Immediately image, keeping the field of view and parameters the same as the baseline image.

For batch processing in calcium imaging data analysis, CNMF-E44 was used. The code for parameters (getmat.m) can be found in the source. The sum of neuron.S, representing all neurons in the view, is recorded as the total number of spikes after processing. The number of active cells is recorded as the number of rows of neuron.S, i.e. the ROI where the spike is detected. The calcium trace overlay plot is presented by CNMFeClustering21. Calcium imaging analysis of microglia was initially conducted by isolating each microglial cell based on the maximum intensity projection of image sequences in the Cell Tracker channel. Subsequently, the Fluo-4 AM sequence image data for each isolated microglia was analyzed. Spikes were defined and quantified using a specific code, which is provided for reference.

Statistical analysis

All statistical analyses in this study were conducted using commercially available software, namely Prism and Excel. Data from replicates are presented as mean ± standard error of the mean (SEM). The methodology for analyzing differences between multiple groups is detailed in the figure legends accompanying each graph. For the creation of graphical representations, Adobe Illustrator 2024 and Biorender were utilized.

Code availability

All original code has been deposited at Zenodo and Github and is publicly available as of the date of publication.